---
title: "MR-Hevo: theoretical derivations"
subtitle: "LD adjustment for summary statistics; GATE score coefficient"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    latex_engine: pdflatex
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(data.table)
library(knitr)
library(kableExtra)
```

\part{Recovering multivariate regression coefficients from univariate summary statistics via the LD correlation matrix}

# Introduction

In genome-wide association studies (GWAS), summary statistics are typically reported as
*marginal* (univariate) regression coefficients $\hat\beta_{u,j}$, one per SNP,
estimated by regressing the trait on each SNP separately.  When a locus contains
$P$ SNPs in linkage disequilibrium (LD), the marginal coefficient for SNP $j$ conflates
its direct effect on the trait with indirect contributions absorbed from correlated
neighbours.

This document derives the formula for recovering *multivariate* (joint) regression
coefficients $\hat{\bm\beta}_m$ from the marginal coefficients, using the LD correlation
matrix among SNPs.  The derivation works with **unstandardized** genotype units
(allele dosage 0/1/2) throughout: both the input ($\hat{\bm\beta}_u$) and the output
($\hat{\bm\beta}_m$) are in units of trait change per additional copy of the effect
allele.  Allele frequencies are assumed available so that the genotype variances can be
computed.

# Notation

Let $n$ be the number of individuals and $P$ the number of SNPs in a locus.

- $\mathbf{g}_j \in \mathbb{R}^n$: vector of allele dosages for SNP $j$
  (centred, but **not** scaled to unit variance).
  Each $g_{ij} \in \{0, 1, 2\}$ minus the mean $2f_j$.
- $\sigma_j^2 = \operatorname{Var}(g_j) = 2f_j(1 - f_j)$: genotype variance for SNP $j$
  with allele frequency $f_j$.
- $\mathbf{D} = \operatorname{diag}(\sigma_1, \dots, \sigma_P)$: diagonal matrix of
  genotype standard deviations.
- $\mathbf{G} = [\mathbf{g}_1 \mid \cdots \mid \mathbf{g}_P]$: $n \times P$
  unstandardized genotype matrix.
- $\mathbf{G}^* = \mathbf{G}\mathbf{D}^{-1}$: genotype matrix with each column scaled
  to unit variance ($\mathbf{G}^* = [\mathbf{g}_1/\sigma_1 \mid \cdots \mid \mathbf{g}_P/\sigma_P]$).
- $\mathbf{C}_g = \mathbf{G}^{*\top}\mathbf{G}^*/n$: $P\times P$ LD **correlation** matrix
  with $[\mathbf{C}_g]_{jj} = 1$ and $[\mathbf{C}_g]_{jk} = r_{jk}$ (the Pearson
  correlation between SNPs $j$ and $k$).
- $\boldsymbol{\Sigma}_g = \mathbf{G}^\top\mathbf{G}/n$: $P\times P$
  **covariance** matrix of the unstandardized genotypes.
- $\mathbf{y} \in \mathbb{R}^n$: centred trait vector.

The covariance and correlation matrices are related by
$$
\boldsymbol{\Sigma}_g = \mathbf{D}\,\mathbf{C}_g\,\mathbf{D}.
\tag{1}
$$

# Multivariate regression: the target

The joint ordinary-least-squares (OLS) estimator $\hat{\bm\beta}_m \in \mathbb{R}^P$ in
$$
\mathbf{y} = \mathbf{G}\bm{\beta}_m + \bm{\varepsilon}
$$
is given by the normal equations
$$
(\mathbf{G}^\top\mathbf{G})\,\hat{\bm\beta}_m = \mathbf{G}^\top\mathbf{y}.
\tag{2}
$$
Dividing both sides by $n$:
$$
\boldsymbol{\Sigma}_g\,\hat{\bm\beta}_m = \frac{1}{n}\mathbf{G}^\top\mathbf{y},
\tag{3}
$$
so $\hat{\bm\beta}_m = \boldsymbol{\Sigma}_g^{-1}(\mathbf{G}^\top\mathbf{y}/n)$ whenever
$\boldsymbol{\Sigma}_g$ is invertible.  The entries $\hat\beta_{m,j}$ are in units of
**trait change per allele** --- the same units as the marginal coefficients.

# Marginal regression coefficients

The univariate OLS estimate from regressing $\mathbf{y}$ on $\mathbf{g}_j$ alone is
$$
\hat\beta_{u,j}
  = \frac{\mathbf{g}_j^\top\mathbf{y}}{\mathbf{g}_j^\top\mathbf{g}_j}
  = \frac{\mathbf{g}_j^\top\mathbf{y}}{n\,\sigma_j^2}.
\tag{4}
$$
Rearranging, the sample covariance of $\mathbf{g}_j$ with $\mathbf{y}$ is
$$
\frac{\mathbf{g}_j^\top\mathbf{y}}{n} = \sigma_j^2\,\hat\beta_{u,j}.
\tag{5}
$$
Stacking over all $j$, with $\mathbf{D}^2 = \operatorname{diag}(\sigma_1^2,\dots,\sigma_P^2)$:
$$
\frac{1}{n}\mathbf{G}^\top\mathbf{y}
  = \mathbf{D}^2\,\hat{\bm\beta}_u.
\tag{6}
$$

# Key identity: multivariate from marginal via the inverse correlation matrix

Substituting (6) into (3) and using the factorization (1):
$$
\boldsymbol{\Sigma}_g\,\hat{\bm\beta}_m = \mathbf{D}^2\,\hat{\bm\beta}_u
\implies
\mathbf{D}\,\mathbf{C}_g\,\mathbf{D}\,\hat{\bm\beta}_m = \mathbf{D}^2\,\hat{\bm\beta}_u.
$$
Premultiplying both sides by $\mathbf{D}^{-1}$:
$$
\mathbf{C}_g\,\mathbf{D}\,\hat{\bm\beta}_m = \mathbf{D}\,\hat{\bm\beta}_u.
$$
Premultiplying both sides by $\mathbf{C}_g^{-1}$:
$$
\mathbf{D}\,\hat{\bm\beta}_m = \mathbf{C}_g^{-1}\,\mathbf{D}\,\hat{\bm\beta}_u.
$$
Finally, premultiplying by $\mathbf{D}^{-1}$:

$$
\boxed{
\hat{\bm\beta}_m = \mathbf{D}^{-1}\,\mathbf{C}_g^{-1}\,\mathbf{D}\,\hat{\bm\beta}_u.
}
\tag{7}
$$

Both $\hat{\bm\beta}_u$ and $\hat{\bm\beta}_m$ are in **units of trait change per allele**;
$\mathbf{D}$ converts between per-allele and per-SD units internally.

## Interpretation via per-SD coefficients

It is instructive to introduce per-SD (standardized) marginal coefficients
$\hat\beta_{u,j}^* = \sigma_j\,\hat\beta_{u,j}$, i.e.\
$\hat{\bm\beta}_u^* = \mathbf{D}\,\hat{\bm\beta}_u$, and per-SD multivariate
coefficients $\hat{\bm\beta}_m^* = \mathbf{D}\,\hat{\bm\beta}_m$.
Then equation (7) decomposes into three steps:

\begin{enumerate}
\item \textbf{Scale} marginal coefficients to per-SD units:
      $\hat{\bm\beta}_u^* = \mathbf{D}\,\hat{\bm\beta}_u$.
\item \textbf{Apply} the inverse correlation matrix:
      $\hat{\bm\beta}_m^* = \mathbf{C}_g^{-1}\,\hat{\bm\beta}_u^*$.
\item \textbf{Rescale} back to per-allele units:
      $\hat{\bm\beta}_m = \mathbf{D}^{-1}\hat{\bm\beta}_m^*$.
\end{enumerate}

Steps 1--2 recover the per-SD multivariate coefficients (the formula for the fully
standardized case); step 3 converts back to the original per-allele units.

## Verification

We can verify equation (7) by checking that $\hat{\bm\beta}_m$ satisfies the normal
equations (3):
$$
\boldsymbol{\Sigma}_g\,\hat{\bm\beta}_m
  = \mathbf{D}\mathbf{C}_g\mathbf{D} \cdot \mathbf{D}^{-1}\mathbf{C}_g^{-1}\mathbf{D}\,\hat{\bm\beta}_u
  = \mathbf{D}\mathbf{C}_g(\mathbf{C}_g^{-1})\mathbf{D}\,\hat{\bm\beta}_u
  = \mathbf{D}^2\hat{\bm\beta}_u
  = \frac{1}{n}\mathbf{G}^\top\mathbf{y}. \quad \checkmark
$$

# Application to GWAS summary statistics

In practice the discovery GWAS provides only per-allele marginal coefficients $\hat\beta_{u,j}$
and their standard errors; individual genotypes are unavailable.  The LD correlation matrix
$\mathbf{C}_g$ is estimated from a reference panel of $n_\text{ref}$ individuals with
similar ancestry (e.g.\ 1000 Genomes EUR, $n_\text{ref}=503$):
$$
\hat{\mathbf{C}}_g = \widehat{\mathbf{G}}^{*\top}\widehat{\mathbf{G}}^*/n_\text{ref},
$$
where $\widehat{\mathbf{G}}^*$ is the reference panel genotype matrix with each column
standardized to unit variance.  The per-SNP SDs $\hat\sigma_j$ are likewise taken from the
reference panel.  Equation (7) then gives
$$
\hat{\bm\beta}_m = \hat{\mathbf{D}}^{-1}\,\hat{\mathbf{C}}_g^{-1}\,\hat{\mathbf{D}}\,\hat{\bm\beta}_u,
\qquad \hat{\mathbf{D}} = \operatorname{diag}(\hat\sigma_1,\dots,\hat\sigma_P).
$$
This approximation is exact when the reference-panel LD matches the discovery-sample LD.

## Scalar instrument construction

For Mendelian randomization, the $P$ per-allele multivariate coefficients at a locus
are combined into a single scalar instrument.  Define the locus score for individual $i$:
$$
S_i = \mathbf{g}_i^\top \hat{\bm\beta}_m,
$$
and the scalar instrument-exposure coefficient
$$
\hat\alpha = \|\hat{\bm\beta}_m\|,
$$
i.e.\ the Euclidean norm of the **per-allele** multivariate coefficient vector.  The
scalar instrument is then
$$
Z_i = \frac{S_i}{\hat\alpha} = \frac{\mathbf{g}_i^\top\hat{\bm\beta}_m}{\|\hat{\bm\beta}_m\|}.
$$

**Justification.** In expectation, $X = \mathbf{g}^\top\bm\beta_m + \varepsilon_X$, so
$$
\operatorname{Cov}(X, Z)
  = \frac{\operatorname{Var}(\mathbf{g}^\top\bm\beta_m)}{\hat\alpha}
  = \frac{\bm\beta_m^\top\boldsymbol\Sigma_g\bm\beta_m}{\hat\alpha},
\qquad
\operatorname{Var}(Z) = \frac{\bm\beta_m^\top\boldsymbol\Sigma_g\bm\beta_m}{\hat\alpha^2}.
$$
The regression coefficient of $X$ on $Z$ (through the origin) is therefore
$$
\frac{\operatorname{Cov}(X,Z)}{\operatorname{Var}(Z)}
  = \frac{\bm\beta_m^\top\boldsymbol\Sigma_g\bm\beta_m/\hat\alpha}
         {\bm\beta_m^\top\boldsymbol\Sigma_g\bm\beta_m/\hat\alpha^2}
  = \hat\alpha,
$$
confirming $\mathbb{E}[X \mid Z] \approx \hat\alpha\,Z$.

## Standard error of $\hat\alpha$

The Fisher information about $\alpha$ from $N$ observations of the instrument-exposure
regression $X = \alpha Z + \varepsilon_X$ is
$$
\mathcal{I}_\alpha = N\,\frac{\operatorname{Var}(Z)}{\operatorname{Var}(\varepsilon_X)},
$$
where $\operatorname{Var}(\varepsilon_X) = \sigma_X^2 - \alpha^2\operatorname{Var}(Z)$
is the residual variance of $X$ after projecting onto $Z$.

**Evaluating $\operatorname{Var}(Z)$.**  With $\hat\alpha = \|\hat{\bm\beta}_m\|$:
$$
\operatorname{Var}(Z) = \frac{\hat{\bm\beta}_m^\top\boldsymbol\Sigma_g\hat{\bm\beta}_m}
                            {\|\hat{\bm\beta}_m\|^2},
$$
estimated empirically from the reference panel as
$\widehat{\operatorname{Var}}(Z) = \widehat{\operatorname{Var}}(S)\,/\,\hat\alpha^2$,
where $\widehat{\operatorname{Var}}(S) = n_\text{ref}^{-1}\sum_i S_i^2$ and
$S_i = \mathbf{g}_i^\top\hat{\bm\beta}_m$ are the reference-panel scores.

Note that
$$
\hat\alpha^2\operatorname{Var}(Z) = \hat{\bm\beta}_m^\top\boldsymbol\Sigma_g\hat{\bm\beta}_m
= \operatorname{Var}(S),
$$
the total variance of the locus score, so the residual variance simplifies to
$\sigma_X^2 - \operatorname{Var}(S)$.

**Estimating $\sigma_X^2$.**  The variance of the exposure $X$ is not directly available
from GWAS summary statistics, but can be recovered from the marginal association of any
single SNP $j$ with $X$.  From the OLS model $X = \beta_{u,j} g_j + \varepsilon_j$,
$$
\operatorname{SE}(\hat\beta_{u,j})^2
  = \frac{\sigma_X^2}{N\,\sigma_j^2}(1 - R_j^2)
  \approx \frac{\sigma_X^2}{N\,\sigma_j^2},
$$
giving
$$
\sigma_X^2 \approx \sigma_j^2\bigl(N\,\operatorname{SE}(\hat\beta_{u,j})^2 + \hat\beta_{u,j}^2\bigr).
\tag{9}
$$
Taking the median of equation (9) over all $P$ SNPs in the locus reduces the influence
of SNPs with unusually large or small marginal associations.

**The standard error.**  Assembling the pieces:
$$
\boxed{
\operatorname{SE}(\hat\alpha) = \frac{1}{\sqrt{\mathcal{I}_\alpha}}
  = \sqrt{\frac{\sigma_X^2 - \widehat{\operatorname{Var}}(S)}
               {N\,\widehat{\operatorname{Var}}(Z)}}.
}
\tag{10}
$$
The $z$-score $\hat\alpha/\operatorname{SE}(\hat\alpha)$ tests whether the locus has
any net association with the exposure.  Because both $\hat\alpha$ and
$\operatorname{SE}(\hat\alpha)$ scale proportionally with $\|\hat{\bm\beta}_m\|$ (as
shown by the formula), the $z$-score is scale-invariant: it is the same whether
$\hat\alpha = \|\hat{\bm\beta}_m\|$ (per-allele) or any positive scalar multiple thereof.

# Instrument-outcome coefficient $\hat\gamma$

## Setup

Let $Y$ denote the outcome (binary or continuous) and let $\hat{\bm\gamma}_u \in\mathbb{R}^P$
be the vector of **per-allele univariate** SNP-outcome coefficients from an outcome GWAS
of $N_Y$ individuals:
$$
\hat\gamma_{u,j} = \frac{\operatorname{Cov}(g_j, Y)}{\operatorname{Var}(g_j)}
= \frac{\operatorname{Cov}(g_j, Y)}{\sigma_j^2}.
\tag{11}
$$
For a **binary** outcome (case-control logistic regression), $\hat\gamma_{u,j}$ is the
per-allele log-odds ratio.  If only a $z$-score $z_j = \hat\gamma_{u,j}/\operatorname{SE}(\hat\gamma_{u,j})$
is provided, the per-allele coefficient is recovered as
$$
\hat\gamma_{u,j} = \frac{z_j}{\sqrt{N_Y\,\sigma_j^2\,p(1-p)}},
\tag{12}
$$
where $p$ is the proportion of cases and $\sigma_j^2 = \operatorname{Var}(g_j)$ is
estimated from the reference panel.

## Multivariate outcome coefficients from univariate summary statistics

By exactly the same argument as in Section 5, applying the inverse LD correlation
matrix recovers the per-allele multivariate outcome coefficients:
$$
\hat{\bm\gamma}_m = \hat{\mathbf{D}}^{-1}\,\hat{\mathbf{C}}_g^{-1}\,\hat{\mathbf{D}}\,\hat{\bm\gamma}_u.
\tag{13}
$$
These are the slopes that $Y$ would receive in a joint regression of $Y$ on all $P$ SNPs
simultaneously.  Define the outcome score for individual $i$ in the reference panel:
$$
T_i = \mathbf{g}_i^\top\hat{\bm\gamma}_m.
$$

## Reference-panel OLS estimator of $\hat\gamma$

With individual-level genotypes available in the reference panel but not in the outcome
GWAS, we estimate $\hat\gamma$ by regressing the outcome score $T$ on the scalar
instrument $Z$ through the origin:
$$
\hat\gamma = \frac{\sum_{i=1}^{n_\text{ref}} T_i Z_i}{\sum_{i=1}^{n_\text{ref}} Z_i^2}.
\tag{14}
$$

**Justification.** In the population, $\mathbb{E}[Y \mid \mathbf{g}] \approx \mathbf{g}^\top\bm\gamma_m = T$,
so (14) approximates $\operatorname{Cov}(Y, Z)/\operatorname{Var}(Z)$, which is the
instrument-outcome regression coefficient.

**Closed form.** Expanding $T_i = \mathbf{g}_i^\top\hat{\bm\gamma}_m$ and
$Z_i = \mathbf{g}_i^\top\hat{\bm\beta}_m/\hat\alpha$:
$$
\sum_i T_i Z_i
  = \frac{1}{\hat\alpha}\sum_i (\mathbf{g}_i^\top\hat{\bm\gamma}_m)(\mathbf{g}_i^\top\hat{\bm\beta}_m)
  = \frac{n_\text{ref}}{\hat\alpha}\,\hat{\bm\gamma}_m^\top\hat{\boldsymbol\Sigma}_g\hat{\bm\beta}_m,
$$
$$
\sum_i Z_i^2
  = \frac{1}{\hat\alpha^2}\sum_i (\mathbf{g}_i^\top\hat{\bm\beta}_m)^2
  = \frac{n_\text{ref}}{\hat\alpha^2}\,\hat{\bm\beta}_m^\top\hat{\boldsymbol\Sigma}_g\hat{\bm\beta}_m
  = \frac{n_\text{ref}}{\hat\alpha^2}\,\widehat{\operatorname{Var}}(S).
$$
Therefore
$$
\hat\gamma
  = \hat\alpha\,\frac{\hat{\bm\gamma}_m^\top\hat{\boldsymbol\Sigma}_g\hat{\bm\beta}_m}
                     {\hat{\bm\beta}_m^\top\hat{\boldsymbol\Sigma}_g\hat{\bm\beta}_m}.
\tag{15}
$$
Using $\hat{\bm\gamma}_m = \hat{\mathbf{D}}^{-1}\hat{\mathbf{C}}_g^{-1}\hat{\mathbf{D}}\hat{\bm\gamma}_u$
and $\hat{\boldsymbol\Sigma}_g = \hat{\mathbf{D}}\hat{\mathbf{C}}_g\hat{\mathbf{D}}$:
$$
\hat{\bm\gamma}_m^\top\hat{\boldsymbol\Sigma}_g\hat{\bm\beta}_m
  = \hat{\bm\gamma}_u^\top\hat{\mathbf{D}}\hat{\mathbf{C}}_g^{-1}\hat{\mathbf{C}}_g\hat{\mathbf{D}}\hat{\bm\beta}_m
  = \hat{\bm\gamma}_u^\top\hat{\mathbf{D}}^2\hat{\bm\beta}_m
  = \sum_{j=1}^P \hat\sigma_j^2\,\hat\gamma_{u,j}\,\hat\beta_{m,j}.
$$
Substituting and writing $\widehat{\operatorname{Var}}(S) = \hat{\bm\beta}_m^\top\hat{\boldsymbol\Sigma}_g\hat{\bm\beta}_m$:
$$
\boxed{
\hat\gamma
  = \hat\alpha\,\frac{\displaystyle\sum_{j=1}^P \hat\sigma_j^2\,\hat\gamma_{u,j}\,\hat\beta_{m,j}}
                     {\widehat{\operatorname{Var}}(S)}.
}
\tag{16}
$$
This shows that $\hat\gamma$ is a weighted sum of the per-allele products
$\hat\gamma_{u,j}\hat\beta_{m,j}$, with weights $\hat\sigma_j^2$ proportional to the
genotype variance of each SNP.

## Standard error of $\hat\gamma$

For a **binary** outcome with proportion $p$ of cases, the Fisher information for
the slope $\gamma$ in a logistic regression on $Z$ is
$$
\mathcal{I}_\gamma = N_Y\,\operatorname{Var}(Z)\,p(1-p),
$$
giving
$$
\boxed{
\operatorname{SE}(\hat\gamma) = \frac{1}{\sqrt{\mathcal{I}_\gamma}}
  = \frac{1}{\sqrt{N_Y\,\widehat{\operatorname{Var}}(Z)\,p(1-p)}}
  = \frac{\hat\alpha}{\sqrt{N_Y\,\widehat{\operatorname{Var}}(S)\,p(1-p)}}.
}
\tag{17}
$$
For a **continuous** outcome with residual variance $\sigma_Y^2$,
$\mathcal{I}_\gamma = N_Y\operatorname{Var}(Z)/\sigma_Y^2$ and
$\operatorname{SE}(\hat\gamma) = \sigma_Y/\sqrt{N_Y\operatorname{Var}(Z)}$.

Note that $\operatorname{SE}(\hat\gamma)$ depends on $\widehat{\operatorname{Var}}(Z) = \widehat{\operatorname{Var}}(S)/\hat\alpha^2$ computed from the reference panel, the outcome sample size $N_Y$, and (for binary outcomes) the case fraction $p$.  It does not require knowledge of $\sigma_Y^2$.

# Handling near-singular LD matrices

When $P \gg n_\text{ref}$ or SNPs are in near-perfect LD, $\hat{\mathbf{C}}_g$ is
ill-conditioned.  The **truncated pseudoinverse** stabilises the computation.

Write the eigendecomposition
$$
\hat{\mathbf{C}}_g = \mathbf{V}\operatorname{diag}(d_1,\dots,d_P)\mathbf{V}^\top,
\qquad d_1 \ge \cdots \ge d_P \ge 0,
$$
and retain only the $K$ eigenvalues with $d_k > \epsilon_0\,d_1$ (default
$\epsilon_0 = 0.01$).  The pseudoinverse is
$$
\hat{\mathbf{C}}_g^{+}
  = \mathbf{V}_K\operatorname{diag}(d_1^{-1},\dots,d_K^{-1})\mathbf{V}_K^\top.
$$
The regularised estimate is then
$$
\hat{\bm\beta}_m = \hat{\mathbf{D}}^{-1}\,\hat{\mathbf{C}}_g^{+}\,\hat{\mathbf{D}}\,\hat{\bm\beta}_u.
\tag{8}
$$
The effective rank $K$ measures the number of independent dimensions of LD variation
in the locus.

# Summary: the complete algorithm

\subsection*{Instrument-exposure coefficients $\hat\alpha$}
\begin{enumerate}
\item \textbf{Input}: per-allele marginal exposure coefficients $\hat\beta_{u,j}$
      and standard errors $\hat s_j$ from the exposure GWAS ($N$ individuals);
      reference-panel SDs $\hat\sigma_j$ (or allele frequencies $\hat f_j$).
\item Form $\hat{\mathbf{D}} = \operatorname{diag}(\hat\sigma_1,\dots,\hat\sigma_P)$;
      compute $\hat{\mathbf{C}}_g$ from the reference panel;
      invert via truncated eigendecomposition (threshold $\epsilon_0$) to get $\hat{\mathbf{C}}_g^+$.
\item Apply eq.\ (8): $\hat{\bm\beta}_m = \hat{\mathbf{D}}^{-1}\hat{\mathbf{C}}_g^+\hat{\mathbf{D}}\,\hat{\bm\beta}_u$.
\item Set $\hat\alpha = \|\hat{\bm\beta}_m\|$ and $Z_i = \mathbf{g}_i^\top\hat{\bm\beta}_m/\hat\alpha$.
\item Estimate $\widehat{\operatorname{Var}}(S)$ from reference-panel scores
      $S_i = \mathbf{g}_i^\top\hat{\bm\beta}_m$; estimate $\hat\sigma_X^2$ via eq.\ (9);
      compute $\operatorname{SE}(\hat\alpha)$ via eq.\ (10).
\end{enumerate}

\subsection*{Instrument-outcome coefficients $\hat\gamma$}
\begin{enumerate}
\item \textbf{Input}: per-allele marginal outcome coefficients $\hat\gamma_{u,j}$
      from the outcome GWAS ($N_Y$ individuals, case fraction $p$ for binary outcomes).
      If only $z$-scores are available, recover $\hat\gamma_{u,j}$ via eq.\ (12).
\item Apply eq.\ (13) using the \emph{same} $\hat{\mathbf{D}}$, $\hat{\mathbf{C}}_g^+$
      as for the exposure:
      $\hat{\bm\gamma}_m = \hat{\mathbf{D}}^{-1}\hat{\mathbf{C}}_g^+\hat{\mathbf{D}}\,\hat{\bm\gamma}_u$.
\item Compute outcome scores $T_i = \mathbf{g}_i^\top\hat{\bm\gamma}_m$ in the reference panel.
\item Estimate $\hat\gamma$ via eq.\ (14) (reference-panel OLS of $T$ on $Z$),
      or equivalently via the closed form eq.\ (16).
\item Compute $\operatorname{SE}(\hat\gamma)$ via eq.\ (17).
\end{enumerate}

The central result, equation (7), states that premultiplying the vector of marginal
per-allele coefficients by $\hat{\mathbf{D}}^{-1}\hat{\mathbf{C}}_g^{-1}\hat{\mathbf{D}}$
--- equivalently, scaling to per-SD units, premultiplying by the inverse correlation
matrix, and rescaling back --- recovers the joint multivariate per-allele coefficients
without requiring individual-level data.

\cleardoublepage

\part{GATE score coefficient from summary-level pQTL statistics}

# Background

The **GATE** (Genetically Aided Trait Estimation) method constructs a
composite genetic score for an exposure by summing the genetically predicted
exposure values over all trans-pQTLs, then tests this GATE score for
association with a disease outcome.
This part derives a formula that computes the log-OR per SD of the GATE
score---and its standard error---directly from the per-locus summary
statistics $(\hat\alpha_k, \hat\gamma_k, \mathrm{SE}(\hat\gamma_k),
\mathrm{sd}_{Z,k})$ that are already available from the MR-Hevo pipeline,
without requiring access to individual-level data.

# Notation

| Symbol | Meaning |
|--------|---------|
| $k$ | Trans-pQTL locus index; $k = 1,\ldots,K$ |
| $j$ | SNP index within locus $k$ |
| $G_{ij}$ | Genotype (0/1/2 copies of effect allele) for individual $i$, SNP $j$ |
| $\hat\beta_{m,kj}$ | Per-allele instrument weight for SNP $j$ in locus $k$ (from UKBB zarr LD) |
| $Z_{ik} = \sum_j G_{ij}\hat\beta_{m,kj}$ | Scalar instrument score for individual $i$ at locus $k$ |
| $\hat\alpha_k$ | Instrument-exposure coefficient: $E[\Delta X \mid \Delta Z_k = 1]$ |
| $\hat\gamma_k$ | Instrument-outcome coefficient: $E[\Delta \log\mathrm{OR}(Y) \mid \Delta Z_k = 1]$ |
| $\mathrm{SE}(\hat\gamma_k)$ | Standard error of $\hat\gamma_k$ (from LD-approximate formula, eq.\ (17)) |
| $\mathrm{sd}_{Z,k}$ | SD of $Z_k$ in the UKBB reference panel (stored as \texttt{sd.Z}) |
| $N, p$ | Total sample size and case fraction in the target dataset |

# The GATE score

The GATE score for individual $i$, summed over all trans-pQTLs, is
$$
G_i = \sum_{k=1}^K \hat\alpha_k\, Z_{ik},
$$
where the cis-pQTL is excluded.
Trans-pQTL loci are in distinct genomic regions and are therefore
approximately uncorrelated, so
$$
\mathrm{Var}_{\mathrm{ref}}(G) = \sum_k \hat\alpha_k^2\, \mathrm{sd}_{Z,k}^2
$$
with no cross-terms.  The standardised GATE score is
$$
G_i^* = G_i \,/\, \mathrm{SD}_\mathrm{ref}(G).
$$

# Score test for $\hat\beta_\mathrm{GATE}$

## Score and information

Let $Y_i \in \{0,1\}$ denote disease status and let $\hat\mu_i^{(0)}$
denote the fitted probability under the null logistic model (intercept +
covariates, no genetic term).  Define
$$
r_i = Y_i - \hat\mu_i^{(0)}, \qquad
w_i = \hat\mu_i^{(0)}(1-\hat\mu_i^{(0)}).
$$
The score and partial Fisher information for a scalar instrument $X$ are
$$
U_X = \mathbf{x}^\top \mathbf{r}, \qquad
V_X = \mathbf{x}^\top \mathbf{W} \mathbf{x}
     - \mathbf{x}^\top \mathbf{W} \mathbf{Z}
       (\mathbf{Z}^\top \mathbf{W} \mathbf{Z})^{-1}
       \mathbf{Z}^\top \mathbf{W} \mathbf{x},
$$
where $\mathbf{Z}$ is the null-model design matrix (intercept + covariates)
and $\mathbf{W} = \mathrm{diag}(w_i)$.
The score estimate is $\hat\beta = U_X / V_X$ with
$\mathrm{SE}(\hat\beta) = 1/\sqrt{V_X}$.

Since the null model includes an intercept and all covariates, the
per-locus score $U_k = \mathbf{z}_k^\top \mathbf{r}$ already adjusts for
their correlations with $Z_k$.

## Combining over trans-pQTLs

For the standardised composite $G^* = G / \mathrm{SD}_G =
(\sum_k \hat\alpha_k Z_k) / \mathrm{SD}_G$:
$$
U_{G^*} = \frac{1}{\mathrm{SD}_G} \sum_k \hat\alpha_k\, U_k, \qquad
V_{G^*} = \frac{1}{\mathrm{SD}_G^2} \sum_k \hat\alpha_k^2\, V_k,
$$
using the independence of instruments across loci ($U_k \perp U_l$,
$V_k \perp V_l$ for $k \neq l$).

The score estimate is
$$
\hat\beta_\mathrm{GATE}
= \frac{U_{G^*}}{V_{G^*}}
= \mathrm{SD}_G \cdot
  \frac{\displaystyle\sum_k \hat\alpha_k\, U_k}
       {\displaystyle\sum_k \hat\alpha_k^2\, V_k}.
$$
Substituting $U_k = \hat\gamma_k V_k$ (the score estimate for locus $k$):

$$
\boxed{
\hat\beta_\mathrm{GATE}
= \mathrm{SD}_G \cdot
  \frac{\displaystyle\sum_k \hat\alpha_k\, \hat\gamma_k\, /\, \mathrm{SE}(\hat\gamma_k)^2}
       {\displaystyle\sum_k \hat\alpha_k^2\, /\, \mathrm{SE}(\hat\gamma_k)^2}
}
\tag{18}
$$

$$
\boxed{
\mathrm{SE}(\hat\beta_\mathrm{GATE})
= \frac{\mathrm{SD}_G}
       {\sqrt{\displaystyle\sum_k \hat\alpha_k^2\, /\, \mathrm{SE}(\hat\gamma_k)^2}}
}
\tag{19}
$$

# Simplification of the standard error

The LD-approximate SE formula (eq.\ (17) from Part~I) gives
$$
\mathrm{SE}(\hat\gamma_k) = \frac{1}{\sqrt{N\, \mathrm{sd}_{Z,k}^2\, p(1-p)}},
\qquad \text{so} \quad
V_k = \frac{1}{\mathrm{SE}(\hat\gamma_k)^2} = N\,\mathrm{sd}_{Z,k}^2\, p(1-p).
$$
Substituting into the denominator of eq.\ (19):
$$
\sum_k \hat\alpha_k^2\, V_k
= N\, p(1-p) \sum_k \hat\alpha_k^2\, \mathrm{sd}_{Z,k}^2
= N\, p(1-p)\, \mathrm{SD}_G^2.
$$
Therefore
$$
\mathrm{SE}(\hat\beta_\mathrm{GATE})
= \frac{\mathrm{SD}_G}{\sqrt{N\, p(1-p)}\, \mathrm{SD}_G}
= \frac{1}{\sqrt{N\, p(1-p)}}.
$$
The SE is determined entirely by the sample size and case fraction, equal to
the SE from a single 1-df binary logistic regression.

# Equivalence to the IVW z-statistic

Substituting $V_k = N\,\mathrm{sd}_{Z,k}^2\,p(1-p)$ and defining
the per-SD(Z) quantities
$\hat\alpha_k^* = \hat\alpha_k\,\mathrm{sd}_{Z,k}$ and
$\hat\gamma_k^* = \hat\gamma_k\,\mathrm{sd}_{Z,k}$ (the values displayed in
the instrument tables), the z-statistic becomes
$$
z_\mathrm{GATE}
= \frac{\hat\beta_\mathrm{GATE}}{\mathrm{SE}(\hat\beta_\mathrm{GATE})}
= \sqrt{N p(1-p)}\cdot
  \frac{\sum_k (\hat\alpha_k^*)^2\, \hat\theta_k}
       {\sqrt{\sum_k (\hat\alpha_k^*)^2}},
$$
where $\hat\theta_k = \hat\gamma_k^* / \hat\alpha_k^*$ is the per-locus
IV-ratio estimate.  This is the IVW z-statistic with weights
$w_k = (\hat\alpha_k^*)^2$ proportional to the reciprocal variance of the
ratio estimator---identical to the MR-Hevo IVW estimator restricted to
trans-pQTLs.

# Comparison: summary-level vs direct regression (PDCD1/T1D)

The summary-level formula is validated against the directly computed
coefficient from logistic regression of T1D on the individual-level GATE
score in the SDRNT1BIO/GS dataset (4,922 cases, 7,452 controls).
For the direct computation, the GATE score is accumulated as
$G_i = \sum_{k,j} G_{ij} \cdot (\hat\alpha_k \hat\beta_{m,kj})$ across all
trans-pQTL SNPs in the individual-level genotype data on \texttt{diabepi}.

The two methods differ in the SD used for standardisation: the summary-level
formula uses $\mathrm{SD}_\mathrm{ref}(G)$ from the UKBB reference panel,
while the direct regression standardises by the empirical SD in the target
GS dataset.  The z-statistics are invariant to this scaling difference;
the ratio $\hat\beta_\mathrm{summary}/\hat\beta_\mathrm{direct}$ estimates
$\mathrm{SD}_\mathrm{ref}(G) / \mathrm{SD}_\mathrm{target}(G)$ and should
be close to 1 for a well-calibrated reference panel.

```{r comparison-table, results='asis'}
if (file.exists("gate_results.rds")) {
    res <- readRDS("gate_results.rds")
    tab <- data.frame(
        Method          = c("Summary-level", "Direct (individual-level)"),
        `$\\hat\\beta$` = sprintf("%.4f", res$beta_GATE),
        SE              = sprintf("%.5f", res$se_GATE),
        z               = sprintf("%.3f",  res$z_GATE),
        p               = formatC(res$p_value, format="g", digits=3),
        `Trans pQTLs`   = as.integer(res$n_trans_pqtl),
        check.names = FALSE
    )
    kable(tab, format="latex", booktabs=TRUE, escape=FALSE,
          caption=paste0(
              "GATE coefficient for T1D ($\\hat\\beta_\\mathrm{GATE}$ = log-OR per SD of GATE score) ",
              "from summary-level formula (eqs.\\ 18--19) vs direct individual-level regression. ",
              "Trans-pQTLs: number of pQTL loci contributing to the GATE score ",
              "(cis-pQTL PDCD1 excluded). ",
              "SE (summary-level) $= 1/\\sqrt{Np(1-p)} \\approx 0.0184$; ",
              "SE (direct) uses the empirical SD of the GATE score in the GS target dataset."),
          label="gate-comparison") |>
        kable_styling(latex_options="hold_position")
} else {
    cat("\\textit{Run \\texttt{compute\\_gate.R} to generate \\texttt{gate\\_results.rds}.}\n")
}
```

The z-statistics from both methods agree within the range expected from LD
mismatch between the UKBB reference panel and the GS target dataset,
analogous to the attenuation seen in the LD-approximate vs direct $\hat\gamma$
comparison in the main report (Pearson $r \approx 0.83$).
