---
title: "Causal effect of PDCD1 protein on type 1 diabetes: MR-Hevo analysis"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: pdflatex
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{caption}
  - \captionsetup[longtable]{width=\linewidth}
  - \usepackage{url}
  - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.pos = "H")
options(knitr.kable.NA = "")
library(data.table)
library(ggplot2)
library(ggrepel)
library(knitr)
library(kableExtra)
devtools::load_all(".", quiet = TRUE)
```

```{r load-data}
## ---- Load protein-coding gene spans (used for NA gene filling) ----
genes_coding <- fread("refGene.txt.gz", header = FALSE,
                      select = c(2L, 3L, 5L, 6L, 13L),
                      col.names = c("accession", "chrom", "txStart", "txEnd", "symbol"))
genes_coding[, chr_g := suppressWarnings(as.integer(sub("^chr", "", chrom)))]
genes_coding <- genes_coding[!is.na(chr_g) & startsWith(accession, "NM_")]
genes_coding <- genes_coding[, .(txStart = min(txStart), txEnd = max(txEnd)),
                               by = .(chr = chr_g, symbol)]

## Fill NA gene entries with protein-coding genes whose txStart or txEnd
## falls within the locus span [locus_start, locus_end].
fill_na_genes <- function(dt) {
    na_idx <- which(is.na(dt$gene))
    if (length(na_idx) == 0L) return(dt)
    for (i in na_idx) {
        chr_i   <- as.integer(dt$chr[i])
        start_i <- dt$locus_start[i]
        end_i   <- dt$locus_end[i]
        hits <- genes_coding[chr == chr_i &
                              ((txStart >= start_i & txStart <= end_i) |
                               (txEnd   >= start_i & txEnd   <= end_i))]
        if (nrow(hits) > 0L)
            dt$gene[i] <- paste(unique(hits$symbol), collapse = ", ")
    }
    dt
}

## Return the qtlname of the chr2 locus nearest to PDCD1 (hg38 midpoint ~241.86 Mb)
find_cis <- function(dt) {
    pdcd1_mid <- 241858000L
    chr2 <- dt[chr == "2"]
    if (nrow(chr2) == 0L) return(character(0))
    chr2$qtlname[which.min(abs(chr2$locus_start - pdcd1_mid))]
}

## ---- Build per-dataset coefficient tables ----
make_coeffs <- function(alpha_rds, gamma_rds, genes_rds) {
    alpha_dt <- readRDS(alpha_rds)
    gamma_dt  <- readRDS(gamma_rds)
    extra_cols <- intersect(c("sd.Z", "n_ld_blocks"), names(alpha_dt))
    alpha_sel  <- c("qtlname", "chr", "locus_start", "locus_end",
                    "n_snps", "eff_rank", "alpha_hat", "se.alpha_hat", extra_cols)
    dt <- merge(
        alpha_dt[, ..alpha_sel],
        gamma_dt[, .(qtlname, gamma_hat, se.gamma_hat)],
        by = "qtlname"
    )
    dt[, chr := as.character(chr)]
    dt <- get_coeffratios(dt, use.delta = TRUE)
    genes <- readRDS(genes_rds)
    dt <- merge(dt, genes, by = "qtlname", all.x = TRUE)
    dt <- fill_na_genes(dt)
    dt[, chr_sort := suppressWarnings(as.integer(chr))]
    dt[chr == "X", chr_sort := 23L]
    dt[chr == "Y", chr_sort := 24L]
    setorder(dt, chr_sort, locus_start)
    dt[, chr_sort := NULL]
    dt
}

## SCALLOP/UKBB analysis commented out
# coeffs_sc <- make_coeffs("alpha_pdcd1_OID00791.rds", "gamma_t1d.rds", "locus_genes.rds")
# cis_sc    <- find_cis(coeffs_sc)
# ivw_sc    <- get_estimatorsMR(coeffs_sc)

coeffs_ukb <- make_coeffs("alpha_pdcd1_UKB.rds",
                            "gamma_t1d_ukb.rds",
                            "locus_genes_ukb.rds")
cis_ukb       <- find_cis(coeffs_ukb)
ivw_ukb       <- get_estimatorsMR(coeffs_ukb)
ivw_ukb_trans <- get_estimatorsMR(coeffs_ukb[qtlname != cis_ukb])

## ---- Load individual-level T1D coefficients ----
indiv_raw <- fread("pdcd1.mr.coeffs.csv")
setnames(indiv_raw,
         c("chrom", "startpos", "endpos", "alpha.hat", "se.alpha.hat"),
         c("chr",   "locus_start", "locus_end", "alpha_hat", "se.alpha_hat"))
indiv_raw[, chr := as.character(chr)]
## gene label: first entry in the comma-separated gene list
indiv_raw[, gene := sub(",.*$", "", qtlname)]
indiv_raw[, chr_sort := suppressWarnings(as.integer(chr))]
indiv_raw[chr == "X", chr_sort := 23L]
indiv_raw[chr == "Y", chr_sort := 24L]
setorder(indiv_raw, chr_sort, locus_start)
indiv_raw[, chr_sort := NULL]

indiv_dt <- get_coeffratios(indiv_raw, use.delta = TRUE)

## ---- User-provided gene label overrides ----
## Keyed by (chr, position range in bp); applied consistently across all three datasets.
gene_overrides <- list(
    list(chr =  1L, lo = 113e6, hi = 115e6, gene = "PTPN22"),
    list(chr =  2L, lo = 240e6, hi = 243e6, gene = "PDCD1"),
    list(chr =  9L, lo =   5e6, hi =   6e6, gene = "CD274"),
    list(chr = 10L, lo =   5e6, hi =   7e6, gene = "IL15RA"),
    list(chr = 12L, lo = 109e6, hi = 113e6, gene = "SH2B3"),
    list(chr = 19L, lo =  48e6, hi =  50e6, gene = "FUT2")
)
apply_gene_overrides <- function(dt) {
    for (ov in gene_overrides) {
        idx <- which(as.integer(dt$chr) == ov$chr &
                     dt$locus_start >= ov$lo &
                     dt$locus_start <= ov$hi)
        if (length(idx)) set(dt, idx, "gene", ov$gene)
    }
    invisible(dt)
}
# apply_gene_overrides(coeffs_sc)  # SCALLOP commented out
apply_gene_overrides(coeffs_ukb)
apply_gene_overrides(indiv_dt)   # indiv_raw and indiv_dt are the same data.table

cis_indiv       <- indiv_dt[qtl_type == "cis-x", scoreid]
ivw_indiv       <- get_estimatorsMR(indiv_dt)
ivw_indiv_trans <- get_estimatorsMR(indiv_dt[!(scoreid %in% cis_indiv)])

## ---- Load direct gamma comparison (LD approximate vs direct individual regression) ----
comp_direct <- readRDS("gamma_comparison_analytic_vs_direct.rds")
r_ad <- cor(comp_direct$gamma_analytic_sdz, comp_direct$gamma_direct,
            use = "complete.obs")
## Add locus position info from coeffs_ukb for foverlaps matching with indiv_dt
comp_direct <- merge(comp_direct,
                     coeffs_ukb[, .(qtlname, chr, locus_start, locus_end)],
                     by.x = "locus_id", by.y = "qtlname", all.x = TRUE)
comp_direct[, chr := as.character(chr)]

## ---- Gen Scotland individual-level gamma ----
coeffs_gs     <- make_coeffs("alpha_pdcd1_UKB.rds",
                              "gamma_t1d_gs.rds",
                              "locus_genes_gs.rds")
cis_gs        <- find_cis(coeffs_gs)
ivw_gs        <- get_estimatorsMR(coeffs_gs)
ivw_gs_trans  <- get_estimatorsMR(coeffs_gs[qtlname != cis_gs])
apply_gene_overrides(coeffs_gs)
```

# Introduction

We estimate the causal effect of circulating PDCD1 (programmed cell death protein 1)
levels on risk of type 1 diabetes (T1D) using Mendelian randomization with the
regularized horseshoe prior (MR-Hevo).

The T1D outcome dataset is the same across all three analyses: cases from the
Scottish T1D Bioresource (SDRNT1BIO) and controls from Generation Scotland.

Results are presented from three analyses, differing in the LD reference panel
used to construct scalar instruments and in whether instrument-outcome
coefficients $\hat\gamma$ are estimated from individual-level genotypes or
from GWAS summary statistics:

1. **Analysis 1 (Zhou et al.)**: $\hat\alpha$ and $\hat\gamma$ from Zhou et al.,
   who used the 1000 Genomes EUR panel (503 samples) as the LD reference.

2. **Analysis 2 (directly computed $\hat\gamma$, UKBB 18M EUR LD)**: $\hat\alpha$
   from UK Biobank Olink proteomics ($N = 33{,}902$); $\hat\gamma$ computed by
   scoring each individual on $Z_i = \sum_j G_{ij}\,\hat\beta_{m,j}$ and regressing
   T1D on $Z_i$ (4,922 cases, 7,452 controls; adjusted for sex, age, and the first
   three principal components). Instrument weights $\hat\beta_m$ from the UKBB EUR
   LD matrix (362,063 samples, 18 million variants).

3. **Analysis 3 (LD approximate $\hat\gamma$, UKBB 18M EUR LD)**: same $\hat\alpha$
   and instrument weights $\hat\beta_m$ as Analysis 2; $\hat\gamma$ estimated from
   per-SNP SNP-outcome coefficients via the LD approximate projection formula,
   using the same UKBB EUR LD matrix.

Scalar genetic instruments are constructed from loci associated with PDCD1 at
$p < 10^{-6}$, excluding the HLA region (chromosome 6: 25--34 Mb).
Each scalar instrument $Z$ is a linear combination of SNPs within the locus,
constructed so that $\mathrm{SD}(Z) = 1$ in the target dataset.
The instrument-exposure coefficient $\hat\alpha$ is therefore the expected change
in the exposure per change of 1 SD in the scalar instrument $Z$;
it can be interpreted directly as the strength of the instrument.
Throughout this report all displayed values of $\hat\alpha$ and $\hat\gamma$ are
in per-SD($Z$) units, obtained by multiplying the per-unit-$Z$ coefficients
(as stored in the intermediate \texttt{.rds}/\texttt{.csv} files) by
$\mathrm{SD}(Z)$.

```{r instruments-chunk, echo=FALSE}
make_table <- function(dt, label, caption_extra = "",
                        has_eff_rank = TRUE, has_sd_z = FALSE,
                        scale_by_sdz = FALSE, small_font = FALSE,
                        has_ld_blocks = FALSE) {
    sdz <- if (scale_by_sdz && "sd.Z" %in% names(dt)) dt$sd.Z else 1
    base_cols <- list(
        `Nearest Gene`         = dt$gene,
        Chr                    = dt$chr,
        `Pos (Mb)`             = round(dt$locus_start / 1e6, 2),
        `Width (Mb)`           = round((dt$locus_end - dt$locus_start) / 1e6, 2),
        `Num SNPs`             = dt$n_snps
    )
    if (has_eff_rank)
        base_cols[["eff rank"]] <- dt$eff_rank
    if (has_sd_z)
        base_cols[["SD($Z$)"]] <- round(dt$sd.Z, 3)
    if (has_ld_blocks)
        base_cols[["LD blks"]] <- dt$n_ld_blocks
    base_cols[["$\\hat\\alpha$ (SE)"]] <- sprintf("%.4f (%.4f)", dt$alpha_hat * sdz, dt$se.alpha_hat * sdz)
    base_cols[["$\\hat\\gamma$ (SE)"]] <- sprintf("%.4f (%.4f)", dt$gamma_hat * sdz, dt$se.gamma_hat * sdz)
    base_cols[["*"]]                   <- ifelse(abs(dt$gamma_hat / dt$se.gamma_hat) > qnorm(0.995), "*", "")
    tab <- as.data.table(base_cols)

    eff_rank_note  <- if (has_eff_rank)  "eff rank: effective rank of LD matrix. " else ""
    sd_z_note      <- if (has_sd_z)      "SD($Z$): standard deviation of scalar instrument in reference panel. " else ""
    ld_blocks_note <- if (has_ld_blocks) "LD blks: number of zarr LD blocks spanned by the locus. " else ""
    k <- kable(tab, format = "latex", booktabs = TRUE, escape = FALSE,
          longtable = TRUE,
          caption = paste0(
              "Instrument coefficients", caption_extra, ". ",
              "Nearest Gene: protein-coding gene(s) nearest to or overlapping the locus. ",
              "Num SNPs: SNPs in locus; ", eff_rank_note, sd_z_note, ld_blocks_note,
              "$\\hat\\alpha$ and $\\hat\\gamma$: instrument-exposure and ",
              "instrument-outcome coefficients per SD of the scalar instrument (SE in parentheses). ",
              "*: $|\\hat\\gamma / \\mathrm{SE}(\\hat\\gamma)| > 2.58$ ",
              "(two-sided $p < 0.01$)."
          ),
          label = label) |>
        kable_styling(latex_options = c("repeat_header", "striped"),
                      font_size = if (small_font) 8 else NULL) |>
        column_spec(1, width = "2.5cm")
    k
}
```

# Data sources

```{r datasources-table, results='asis', echo=FALSE}
ds <- data.frame(
  File = c(
    "\\path{PDCD1_Q15116_OID21396_v1_Oncology.tar}",
    "\\path{z\\_scores\\_gs.rds}",
    "\\path{pdcd1.mr.coeffs.csv}",
    "\\path{LD_Eur_18mvariants/int8/} (genoscores)",
    "\\path{GS_T1biochrall_filtered_rsmapped.bed/.bim/.fam} (diabepi)",
    "\\path{PHENO_T1DM_GST1B_AS_NOREL_NOMODY_NOPT2_NOCPEPAB_NONEO_NORELPCA.samples} (diabepi)",
    "\\path{refpop/kg.2020.hg38.eur.bim}",
    "\\path{refGene.txt.gz}"
  ),
  Description = c(
    "SNP--protein GWAS summary statistics for PDCD1 (UniProt Q15116; Olink Oncology panel, assay OID21396); $N = 33{,}902$ UK Biobank EUR participants",
    "Per-SNP T1D SNP-outcome z-statistics (Wald) computed from individual-level SDRNT1BIO/GS genotypes by logistic regression in \\texttt{run\\_gwas\\_loci.R}; 4,922 cases, 7,452 controls; aligned to \\texttt{bim\\_a1} allele convention",
    "Scalar $\\hat\\alpha$ and $\\hat\\gamma$ coefficients from Analysis 1 (Zhou et al.), with 1000 Genomes EUR LD",
    "Block-LD matrix; UKBB EUR participants; $N = 362{,}063$; $\\sim$18M variants; GRCh37; int8",
    "Individual-level T1D genotypes (SDRNT1BIO cases + GS controls); 25,744 individuals; 7.8M SNPs; GRCh37; plink BED format",
    "T1D case-control phenotype file; 4,922 cases, 7,452 controls; includes sex, age, PC1--3 covariates; Oxford .samples format",
    "Variant index (rsid, chr, pos hg38) for 1000 Genomes EUR panel; 503 samples; used for (chr, pos) $\\to$ rsid lookup",
    "RefSeq protein-coding gene annotations (NM\\_ accessions); used to label loci with the nearest gene"
  ),
  Source = c(
    "UK Biobank Pharma Proteomics Project (UKB-PPP); Sun et al.\\ (2023) \\textit{Nature} 622:329",
    "Derived from GS plink files and phenotype file listed above; output of \\texttt{compute\\_gamma\\_indiv.R}",
    "Computed by Zhou et al.\\ from UKB-PPP Olink data and SDRNT1BIO/GS individual-level data",
    "UK Biobank; computed using magenpy",
    "Scottish Diabetes Research Network Type 1 Bioresource (SDRNT1BIO) and Generation Scotland (GS)",
    "SDRNT1BIO/GS; quality-controlled for relatedness, non-T1D diabetes, and population structure",
    "1000 Genomes Project phase 3, GRCh38 liftover",
    "UCSC Genome Browser, hg38 assembly"
  ),
  stringsAsFactors = FALSE
)
kable(ds, format = "latex", booktabs = TRUE, longtable = TRUE, escape = FALSE,
      caption = "Source data files used in this analysis.",
      label  = "datasources",
      col.names = c("File", "Description", "Source")) |>
  kable_styling(latex_options = c("repeat_header", "striped"), font_size = 8) |>
  column_spec(1, width = "3.5cm") |>
  column_spec(2, width = "6.5cm") |>
  column_spec(3, width = "4.5cm")
```

# Analysis 1 (Zhou et al.): individual-level $\hat\gamma$, 1000 Genomes EUR LD

`r nrow(indiv_dt)` scalar instruments; $\hat\alpha$ and $\hat\gamma$ from Zhou et al.,
computed from individual-level UK Biobank Olink proteomics data and SDRNT1BIO/GS
T1D genotype--phenotype data, using the 1000 Genomes EUR panel for LD.

\newpage

```{r instruments-table-indiv, results='asis'}
make_table(indiv_dt, label = "instruments-indiv",
           caption_extra = " --- Analysis 1 (Zhou et al.): 1000 Genomes EUR LD",
           has_eff_rank = FALSE)
```

\clearpage

```{r scatter-indiv, fig.cap=paste0("Analysis 1 (Zhou et al.), 1000 Genomes EUR LD. $\\hat{\\gamma}$ vs $\\hat{\\alpha}$ (per SD of scalar instrument) for ", nrow(indiv_dt), " instruments. IVW slope estimated from trans-pQTLs only: $\\hat\\theta = ", round(ivw_indiv_trans$Estimate, 3), "$ (SE $", round(ivw_indiv_trans$SE, 3), "$, $p = ", ivw_indiv_trans$pvalue.formatted, "$). Red dashed line: IVW slope (trans only). Red point: PDCD1 cis-pQTL."), fig.width = 7, fig.height = 6}
plot_iv_estimates(
    alpha_hat    = indiv_dt$alpha_hat,
    se.alpha_hat = indiv_dt$se.alpha_hat,
    gamma_hat    = indiv_dt$gamma_hat,
    se.gamma_hat = indiv_dt$se.gamma_hat,
    theta        = ivw_indiv_trans$Estimate,
    qtlname      = indiv_dt$gene,
    highlight    = indiv_dt$scoreid %in% cis_indiv
)
```

# Analysis 2: individual-level $\hat\gamma$ from SDRNT1BIO/GS genotypes, UKBB 18M EUR LD

`r nrow(coeffs_gs)` scalar instruments; $\hat\alpha$ from UK Biobank Olink
proteomics ($N = 33{,}902$). $\hat\gamma$ estimated from individual-level
SDRNT1BIO/GS genotypes (4,922 cases, 7,452 controls), adjusted for sex, age, and
the first three genetic principal components.  Instrument weights $\hat\beta_m$
derived from the UKBB EUR LD matrix (362,063 samples, 18 million variants).

```{r scatter-gs, fig.cap=paste0("Analysis 2: directly computed $\\hat{\\gamma}$ (individual-level logistic regression, UKBB 18M EUR LD instrument weights) vs $\\hat{\\alpha}$ (per SD of scalar instrument). IVW slope estimated from trans-pQTLs only: $\\hat\\theta = ", round(ivw_direct_trans$Estimate, 3), "$ (SE $", round(ivw_direct_trans$SE, 3), "$, $p = ", ivw_direct_trans$pvalue.formatted, "$). Red dashed line: IVW slope (trans only). Red point: PDCD1 cis-pQTL."), fig.width = 7, fig.height = 6}
## Merge direct gamma with alpha info; gamma_direct is per-SD(Z_target) ≈ per-SD(Z_ref)
d_sc <- merge(
    comp_direct[, .(locus_id, gene, gamma_direct, se_gamma_direct)],
    coeffs_ukb[, .(qtlname, alpha_hat, se.alpha_hat, sd.Z)],
    by.x = "locus_id", by.y = "qtlname"
)
## IVW (trans only: exclude cis-pQTL)
d_sc[, theta_IV    := gamma_direct / (alpha_hat * sd.Z)]
d_sc[, se.theta_IV := se_gamma_direct / (alpha_hat * sd.Z)]
d_sc[, inv.var     := se.theta_IV^-2]
is_cis_direct  <- d_sc$locus_id == cis_gs
ivw_direct_all   <- d_sc[, .(Estimate = sum(theta_IV * inv.var) / sum(inv.var),
                               SE       = sqrt(1 / sum(inv.var)))]
ivw_direct_trans <- d_sc[!is_cis_direct,
                          .(Estimate = sum(theta_IV * inv.var) / sum(inv.var),
                            SE       = sqrt(1 / sum(inv.var)))]
ivw_direct_trans[, z := Estimate / SE]
ivw_direct_trans[, pvalue.formatted := format.z.aspvalue(z)]

plot_iv_estimates(
    alpha_hat    = d_sc$alpha_hat    * d_sc$sd.Z,
    se.alpha_hat = d_sc$se.alpha_hat * d_sc$sd.Z,
    gamma_hat    = d_sc$gamma_direct,
    se.gamma_hat = d_sc$se_gamma_direct,
    theta        = ivw_direct_trans$Estimate,
    qtlname      = sub(",.*$", "", d_sc$gene),
    highlight    = is_cis_direct
)
```

# Analysis 3: LD approximate $\hat\gamma$ from SNP-outcome coefficients, UKBB 18M EUR LD

`r nrow(coeffs_ukb)` scalar instruments; $\hat\alpha$ from UK Biobank Olink
proteomics ($N = 33{,}902$). $\hat\gamma$ estimated from per-SNP T1D SNP-outcome
coefficients (from SDRNT1BIO/GS individual-level logistic regression, 4,922 cases,
7,452 controls), LD-adjusted using the UKBB EUR LD matrix (362,063 samples, 18
million variants).  Same instrument weights $\hat\beta_m$ as Analysis 2.

\newpage

```{r instruments-table-ukb, results='asis'}
make_table(coeffs_ukb, label = "instruments-ukb",
           caption_extra = " --- Analysis 3: LD approximate SNP-outcome coefficients, UKBB 18M EUR LD",
           has_sd_z    = "sd.Z" %in% names(coeffs_ukb),
           scale_by_sdz  = TRUE,
           small_font    = TRUE,
           has_ld_blocks = "n_ld_blocks" %in% names(coeffs_ukb))
```

\clearpage

```{r scatter-ukb, fig.cap=paste0("Analysis 3: LD approximate $\\hat{\\gamma}$ (SNP-outcome coefficients + UKBB 18M EUR LD projection formula) vs $\\hat{\\alpha}$ (per SD of scalar instrument) for ", nrow(coeffs_ukb), " instruments. IVW slope estimated from trans-pQTLs only: $\\hat\\theta = ", round(ivw_ukb_trans$Estimate, 3), "$ (SE $", round(ivw_ukb_trans$SE, 3), "$, $p = ", ivw_ukb_trans$pvalue.formatted, "$). Red dashed line: IVW slope (trans only). Red point: PDCD1 cis-pQTL."), fig.width = 7, fig.height = 6}
sdz <- coeffs_ukb$sd.Z
plot_iv_estimates(
    alpha_hat    = coeffs_ukb$alpha_hat    * sdz,
    se.alpha_hat = coeffs_ukb$se.alpha_hat * sdz,
    gamma_hat    = coeffs_ukb$gamma_hat    * sdz,
    se.gamma_hat = coeffs_ukb$se.gamma_hat * sdz,
    theta        = ivw_ukb_trans$Estimate,
    qtlname      = sub(",.*$", "", coeffs_ukb$gene),
    highlight    = coeffs_ukb$qtlname == cis_ukb
)
```

# Comparison of instrument-exposure ($\hat\alpha$) coefficients: Analysis 1 (1000G LD) vs Analyses 2 and 3 (UKBB 18M LD)

```{r alpha-compare, fig.cap="Instrument-exposure coefficient $\\hat{\\alpha}$ (per SD of scalar instrument): Analysis 1 / Zhou et al.\\ with 1000 Genomes EUR LD (x-axis) vs Analysis 2 with UKBB 18M EUR LD (y-axis). Loci matched by chromosomal overlap. Labelled points are outliers from the identity line ($|y-x| > 2\\,\\mathrm{SD}$). Grey dashed line: identity ($y=x$). Blue dotted line: OLS regression through the origin.", fig.width = 6, fig.height = 6}
dt_indiv <- indiv_dt[,  .(gene = gene, chr, lo = locus_start, hi = locus_end,
                           alpha_indiv = alpha_hat)]
dt_ukb   <- coeffs_ukb[, .(chr, lo = locus_start, hi = locus_end,
                             alpha_ukb   = alpha_hat * sd.Z,
                             n_ld_blocks = n_ld_blocks)]
setkeyv(dt_ukb,   c("chr", "lo", "hi"))
setkeyv(dt_indiv, c("chr", "lo", "hi"))
alpha_comp <- foverlaps(dt_indiv, dt_ukb, nomatch = 0L)

slope_ols <- coef(lm(alpha_ukb ~ 0 + alpha_indiv, data = alpha_comp))["alpha_indiv"]
ax_breaks <- pretty(c(0, max(alpha_comp$alpha_indiv, alpha_comp$alpha_ukb, na.rm = TRUE)))

alpha_comp[, resid_id := alpha_ukb - alpha_indiv]
alpha_comp[, outlier_label := ifelse(
    abs(resid_id) > 2 * sd(resid_id, na.rm = TRUE), gene, "")]

ggplot(alpha_comp, aes(x = alpha_indiv, y = alpha_ukb)) +
    geom_hline(yintercept = 0, linewidth = 0.4) +
    geom_vline(xintercept = 0, linewidth = 0.4) +
    geom_abline(slope = 1,         intercept = 0, linetype = "dashed", colour = "grey50") +
    geom_abline(slope = slope_ols, intercept = 0, linetype = "dotted", colour = "blue") +
    geom_point() +
    geom_text_repel(aes(label = outlier_label), size = 3, max.overlaps = Inf) +
    scale_x_continuous(breaks = ax_breaks, limits = range(ax_breaks)) +
    scale_y_continuous(breaks = ax_breaks, limits = range(ax_breaks)) +
    labs(x = expression(hat(alpha)[indiv]),
         y = expression(hat(alpha)[UKB])) +
    coord_fixed() +
    theme_bw()
```

# Comparison of directly computed $\hat\gamma$: Zhou et al.\ (1000G LD) vs new direct (UKBB 18M LD)

Both analyses use individual-level SDRNT1BIO/GS genotypes to compute the scalar
instrument score $Z_i = \sum_j G_{ij}\,\hat\beta_{m,j}$ for each individual and
regress T1D on $Z_i$.  They differ only in the LD reference panel used to derive
the instrument weights $\hat\beta_m$: 1000 Genomes EUR (503 samples, Analysis 1)
vs UKBB EUR 18M variant LD matrix (362,063 samples, Analysis 2).

```{r gamma-direct-direct, fig.cap="Both axes show directly computed $\\hat{\\gamma}$ from individual-level logistic regression of T1D on the scalar instrument score (per SD, log-odds). x-axis: Zhou et al.\\ Analysis 1 using 1000 Genomes EUR LD weights. y-axis: new analysis using UKBB 18M EUR LD weights (Analysis 2). Both analyses use the same SDRNT1BIO/GS individual-level genotypes; they differ only in the LD reference panel used to derive instrument weights $\\hat\\beta_m$. Loci matched by chromosomal overlap. The six largest outliers from the identity line are labelled. Grey dashed line: identity ($y=x$).", fig.width = 6, fig.height = 6}
dt_zhou <- indiv_dt[, .(gene = gene, chr, lo = locus_start, hi = locus_end,
                         gamma_zhou = gamma_hat)]
dt_new  <- comp_direct[!is.na(locus_start),
                        .(chr, lo = locus_start, hi = locus_end,
                          gamma_new_direct = gamma_direct)]
setkeyv(dt_new,  c("chr", "lo", "hi"))
setkeyv(dt_zhou, c("chr", "lo", "hi"))
dd_comp <- foverlaps(dt_zhou, dt_new, nomatch = 0L)

ax_brk_dd <- pretty(range(c(dd_comp$gamma_new_direct, dd_comp$gamma_zhou), na.rm = TRUE))

dd_comp[, resid_id := gamma_new_direct - gamma_zhou]
top6_genes <- dd_comp[order(-abs(resid_id))][seq_len(min(6L, .N)), gene]
dd_comp[, outlier_label := ifelse(gene %in% top6_genes, gene, "")]

ggplot(dd_comp, aes(x = gamma_zhou, y = gamma_new_direct)) +
    geom_hline(yintercept = 0, linewidth = 0.4) +
    geom_vline(xintercept = 0, linewidth = 0.4) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
    geom_point() +
    geom_text_repel(aes(label = outlier_label), size = 3, max.overlaps = Inf) +
    scale_x_continuous(breaks = ax_brk_dd, limits = range(ax_brk_dd)) +
    scale_y_continuous(breaks = ax_brk_dd, limits = range(ax_brk_dd)) +
    labs(x = expression(hat(gamma)["Zhou et al., 1000G LD"]),
         y = expression(hat(gamma)["direct, UKBB 18M LD"])) +
    coord_fixed() +
    theme_bw()
```

# Comparison of $\hat\gamma$: direct individual-level regression vs LD approximate

The direct method (Analysis 2, x-axis, ground truth) regresses T1D on individual
instrument scores using UKBB EUR LD weights.
The LD approximate method (Analysis 3, y-axis) estimates $\hat\gamma$ from the
same per-SNP SNP-outcome coefficients via the LD approximate projection formula.
Both use identical instrument weights $\hat\beta_m$.
Attenuation of the LD approximate estimate relative to the direct estimate reflects
mismatch between the UKBB reference panel LD and the true LD in the GS target
dataset.
The Pearson correlation is $r = `r round(r_ad, 3)`$.

```{r gamma-analytic-vs-direct, fig.cap=paste0("Direct individual-level regression $\\hat{\\gamma}$ (x-axis, ground truth) vs LD approximate $\\hat{\\gamma}$ (y-axis). Both per SD of scalar instrument, log-odds units; same UKBB 18M EUR LD weights used for both. Pearson $r = ", round(r_ad, 3), "$. The six largest outliers from the identity line are labelled. Grey dashed: identity ($y = x$)."), fig.width = 6, fig.height = 6}
comp_direct[, resid_id := gamma_analytic_sdz - gamma_direct]
top6_genes_ad <- comp_direct[order(-abs(resid_id))][seq_len(min(6L, .N)), gene]
comp_direct[, outlier_label := ifelse(gene %in% top6_genes_ad, gene, "")]

ax_brk_ad <- pretty(range(c(comp_direct$gamma_direct,
                             comp_direct$gamma_analytic_sdz), na.rm = TRUE))

ggplot(comp_direct, aes(x = gamma_direct, y = gamma_analytic_sdz)) +
    geom_hline(yintercept = 0, linewidth = 0.4) +
    geom_vline(xintercept = 0, linewidth = 0.4) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
    geom_point() +
    geom_text_repel(aes(label = outlier_label), size = 3, max.overlaps = Inf) +
    scale_x_continuous(breaks = ax_brk_ad, limits = range(ax_brk_ad)) +
    scale_y_continuous(breaks = ax_brk_ad, limits = range(ax_brk_ad)) +
    labs(x = expression(hat(gamma)["direct (individual-level)"]),
         y = expression(hat(gamma)["LD approximate"])) +
    coord_fixed() +
    theme_bw()
```



<!-- SCALLOP/UKBB section commented out
# SCALLOP/UKBB --- summary-level T1D outcome

```{r instruments-table-sc, eval=FALSE, include=FALSE}
make_table(coeffs_sc, label = "instruments-sc",
           caption_extra = " --- SCALLOP/UKBB SNP--protein coefficients, summary-level T1D")
```

```{r scatter-sc, eval=FALSE, include=FALSE}
plot_iv_estimates(
    alpha_hat    = coeffs_sc$alpha_hat,
    se.alpha_hat = coeffs_sc$se.alpha_hat,
    gamma_hat    = coeffs_sc$gamma_hat,
    se.gamma_hat = coeffs_sc$se.gamma_hat,
    theta        = ivw_sc$Estimate,
    qtlname      = coeffs_sc$gene,
    highlight    = coeffs_sc$qtlname == cis_sc
)
```
-->

```{r mrhevo-results, results='asis'}
if (!file.exists("mrhevo_pdcd1_t1d_fit.rds")) {
    cat("\\textit{MR-Hevo posterior results not yet available.",
        "Run \\texttt{run\\_mrhevo\\_pdcd1\\_t1d.R} to generate",
        "\\texttt{mrhevo\\_pdcd1\\_t1d\\_fit.rds}.}\n")
}
```
