---
title: "Causal effect of PDCD1 protein on type 1 diabetes: MR-Hevo analysis"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: pdflatex
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{caption}
  - \captionsetup[longtable]{width=\linewidth}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center")
options(knitr.kable.NA = "")
library(data.table)
library(ggplot2)
library(ggrepel)
library(knitr)
library(kableExtra)
devtools::load_all(".", quiet = TRUE)
```

```{r load-data}
## ---- Load protein-coding gene spans (used for NA gene filling) ----
genes_coding <- fread("refGene.txt.gz", header = FALSE,
                      select = c(2L, 3L, 5L, 6L, 13L),
                      col.names = c("accession", "chrom", "txStart", "txEnd", "symbol"))
genes_coding[, chr_g := suppressWarnings(as.integer(sub("^chr", "", chrom)))]
genes_coding <- genes_coding[!is.na(chr_g) & startsWith(accession, "NM_")]
genes_coding <- genes_coding[, .(txStart = min(txStart), txEnd = max(txEnd)),
                               by = .(chr = chr_g, symbol)]

## Fill NA gene entries with protein-coding genes whose txStart or txEnd
## falls within the locus span [locus_start, locus_end].
fill_na_genes <- function(dt) {
    na_idx <- which(is.na(dt$gene))
    if (length(na_idx) == 0L) return(dt)
    for (i in na_idx) {
        chr_i   <- as.integer(dt$chr[i])
        start_i <- dt$locus_start[i]
        end_i   <- dt$locus_end[i]
        hits <- genes_coding[chr == chr_i &
                              ((txStart >= start_i & txStart <= end_i) |
                               (txEnd   >= start_i & txEnd   <= end_i))]
        if (nrow(hits) > 0L)
            dt$gene[i] <- paste(unique(hits$symbol), collapse = ", ")
    }
    dt
}

## Return the qtlname of the chr2 locus nearest to PDCD1 (hg38 midpoint ~241.86 Mb)
find_cis <- function(dt) {
    pdcd1_mid <- 241858000L
    chr2 <- dt[chr == "2"]
    if (nrow(chr2) == 0L) return(character(0))
    chr2$qtlname[which.min(abs(chr2$locus_start - pdcd1_mid))]
}

## ---- Build per-dataset coefficient tables ----
make_coeffs <- function(alpha_rds, gamma_rds, genes_rds) {
    alpha_dt <- readRDS(alpha_rds)
    gamma_dt  <- readRDS(gamma_rds)
    extra_cols <- intersect(c("sd.Z"), names(alpha_dt))
    alpha_sel  <- c("qtlname", "chr", "locus_start", "locus_end",
                    "n_snps", "eff_rank", "alpha_hat", "se.alpha_hat", extra_cols)
    dt <- merge(
        alpha_dt[, ..alpha_sel],
        gamma_dt[, .(qtlname, gamma_hat, se.gamma_hat)],
        by = "qtlname"
    )
    dt[, chr := as.character(chr)]
    dt <- get_coeffratios(dt, use.delta = TRUE)
    genes <- readRDS(genes_rds)
    dt <- merge(dt, genes, by = "qtlname", all.x = TRUE)
    dt <- fill_na_genes(dt)
    dt[, chr_sort := suppressWarnings(as.integer(chr))]
    dt[chr == "X", chr_sort := 23L]
    dt[chr == "Y", chr_sort := 24L]
    setorder(dt, chr_sort, locus_start)
    dt[, chr_sort := NULL]
    dt
}

## SCALLOP/UKBB analysis commented out
# coeffs_sc <- make_coeffs("alpha_pdcd1_OID00791.rds", "gamma_t1d.rds", "locus_genes.rds")
# cis_sc    <- find_cis(coeffs_sc)
# ivw_sc    <- get_estimatorsMR(coeffs_sc)

coeffs_ukb <- make_coeffs("alpha_pdcd1_UKB.rds",
                            "gamma_t1d_ukb.rds",
                            "locus_genes_ukb.rds")
cis_ukb       <- find_cis(coeffs_ukb)
ivw_ukb       <- get_estimatorsMR(coeffs_ukb)
ivw_ukb_trans <- get_estimatorsMR(coeffs_ukb[qtlname != cis_ukb])

## ---- Load individual-level T1D coefficients ----
indiv_raw <- fread("pdcd1.mr.coeffs.csv")
setnames(indiv_raw,
         c("chrom", "startpos", "endpos", "alpha.hat", "se.alpha.hat"),
         c("chr",   "locus_start", "locus_end", "alpha_hat", "se.alpha_hat"))
indiv_raw[, chr := as.character(chr)]
## gene label: first entry in the comma-separated gene list
indiv_raw[, gene := sub(",.*$", "", qtlname)]
indiv_raw[, chr_sort := suppressWarnings(as.integer(chr))]
indiv_raw[chr == "X", chr_sort := 23L]
indiv_raw[chr == "Y", chr_sort := 24L]
setorder(indiv_raw, chr_sort, locus_start)
indiv_raw[, chr_sort := NULL]

indiv_dt <- get_coeffratios(indiv_raw, use.delta = TRUE)

## ---- User-provided gene label overrides ----
## Keyed by (chr, position range in bp); applied consistently across all three datasets.
gene_overrides <- list(
    list(chr =  1L, lo = 113e6, hi = 115e6, gene = "PTPN22"),
    list(chr =  2L, lo = 240e6, hi = 243e6, gene = "PDCD1"),
    list(chr =  9L, lo =   5e6, hi =   6e6, gene = "CD274"),
    list(chr = 10L, lo =   5e6, hi =   7e6, gene = "IL15RA"),
    list(chr = 12L, lo = 109e6, hi = 113e6, gene = "SH2B3"),
    list(chr = 19L, lo =  48e6, hi =  50e6, gene = "FUT2")
)
apply_gene_overrides <- function(dt) {
    for (ov in gene_overrides) {
        idx <- which(as.integer(dt$chr) == ov$chr &
                     dt$locus_start >= ov$lo &
                     dt$locus_start <= ov$hi)
        if (length(idx)) set(dt, idx, "gene", ov$gene)
    }
    invisible(dt)
}
# apply_gene_overrides(coeffs_sc)  # SCALLOP commented out
apply_gene_overrides(coeffs_ukb)
apply_gene_overrides(indiv_dt)   # indiv_raw and indiv_dt are the same data.table

cis_indiv     <- indiv_dt[qtl_type == "cis-x", scoreid]
ivw_indiv     <- get_estimatorsMR(indiv_dt)
ivw_indiv_trans <- get_estimatorsMR(indiv_dt[!(scoreid %in% cis_indiv)])
```

# Introduction

We estimate the causal effect of circulating PDCD1 (programmed cell death protein 1)
levels on risk of type 1 diabetes (T1D) using Mendelian randomization with the
regularized horseshoe prior (MR-Hevo).

Results are presented from two analyses:

1. **Individual-level T1D outcome (Zhou et al.)**: instrument-exposure coefficients
   from the UK Biobank Olink proteomics study ($N = 33{,}902$); instrument-outcome
   coefficients estimated from individual-level genotype--phenotype data.

2. **Summary-level T1D outcome, UK Biobank SNP--protein coefficients**: both sets of
   coefficients from summary statistics, with SNP--protein associations from the
   UK Biobank Olink study ($N = 33{,}902$) and SNP--T1D associations from
   Iakovliev et al.\ (Am J Hum Genet 2023; 4,964 cases, 7,497 controls).
   LD matrices from the UK Biobank EUR panel (362,063 samples, 18 million variants)
   were used to construct scalar instruments.

Scalar genetic instruments are constructed from loci associated with PDCD1 at
$p < 10^{-6}$, excluding the HLA region (chromosome 6: 25--34 Mb).
Each scalar instrument $Z$ is a linear combination of SNPs within the locus,
constructed so that $\mathrm{SD}(Z) = 1$ in the target dataset.
The instrument-exposure coefficient $\hat\alpha$ is therefore the expected change
in the exposure per change of 1 SD in the scalar instrument $Z$;
it can be interpreted directly as the strength of the instrument.
Throughout this report all displayed values of $\hat\alpha$ and $\hat\gamma$ are
in per-SD($Z$) units, obtained by multiplying the per-unit-$Z$ coefficients
(as stored in the intermediate \texttt{.rds}/\texttt{.csv} files) by
$\mathrm{SD}(Z)$.

```{r instruments-chunk, echo=FALSE}
make_table <- function(dt, label, caption_extra = "",
                        has_eff_rank = TRUE, has_sd_z = FALSE,
                        scale_by_sdz = FALSE, small_font = FALSE) {
    sdz <- if (scale_by_sdz && "sd.Z" %in% names(dt)) dt$sd.Z else 1
    base_cols <- list(
        `Nearest Gene`         = dt$gene,
        Chr                    = dt$chr,
        `Pos (Mb)`             = round(dt$locus_start / 1e6, 2),
        `Width (Mb)`           = round((dt$locus_end - dt$locus_start) / 1e6, 2),
        `Num SNPs`             = dt$n_snps
    )
    if (has_eff_rank)
        base_cols[["eff rank"]] <- dt$eff_rank
    if (has_sd_z)
        base_cols[["SD($Z$)"]] <- round(dt$sd.Z, 3)
    base_cols[["$\\hat\\alpha$ (SE)"]] <- sprintf("%.4f (%.4f)", dt$alpha_hat * sdz, dt$se.alpha_hat * sdz)
    base_cols[["$\\hat\\gamma$ (SE)"]] <- sprintf("%.4f (%.4f)", dt$gamma_hat * sdz, dt$se.gamma_hat * sdz)
    base_cols[["*"]]                   <- ifelse(abs(dt$gamma_hat / dt$se.gamma_hat) > qnorm(0.995), "*", "")
    tab <- as.data.table(base_cols)

    eff_rank_note <- if (has_eff_rank) "eff rank: effective rank of LD matrix. " else ""
    sd_z_note     <- if (has_sd_z) "SD($Z$): standard deviation of scalar instrument in reference panel. " else ""
    k <- kable(tab, format = "latex", booktabs = TRUE, escape = FALSE,
          longtable = TRUE,
          caption = paste0(
              "Instrument coefficients", caption_extra, ". ",
              "Nearest Gene: protein-coding gene(s) nearest to or overlapping the locus. ",
              "Num SNPs: SNPs in locus; ", eff_rank_note, sd_z_note,
              "$\\hat\\alpha$ and $\\hat\\gamma$: instrument-exposure and ",
              "instrument-outcome coefficients per SD of the scalar instrument (SE in parentheses). ",
              "*: $|\\hat\\gamma / \\mathrm{SE}(\\hat\\gamma)| > 2.58$ ",
              "(two-sided $p < 0.01$)."
          ),
          label = label) |>
        kable_styling(latex_options = c("repeat_header", "striped"),
                      font_size = if (small_font) 8 else NULL) |>
        column_spec(1, width = "2.5cm")
    k
}
```

# Data sources

The following source data files are used in this analysis.

| File | Description | Source |
|------|-------------|--------|
| `PDCD1_Q15116_OID21396_v1_Oncology.tar` | SNP--protein GWAS summary statistics for PDCD1 (UniProt Q15116; Olink Oncology panel, assay OID21396); $N = 33{,}902$ UK Biobank participants of European ancestry | UK Biobank Pharma Proteomics Project (UKB-PPP); Sun et al.\ (2023) *Nature* 622:329--338 |
| `gwasresults.RData.gz` | Type 1 diabetes GWAS summary statistics; 4,964 cases and 7,497 controls | Iakovliev et al.\ (2023) *Am J Hum Genet* 110:1718--1732 |
| `pdcd1.mr.coeffs.csv` | Scalar instrument--exposure ($\hat\alpha$) and instrument--outcome ($\hat\gamma$) coefficients from the individual-level MR analysis, with scalar instruments constructed using LD from the 1000 Genomes EUR panel | Computed by Zhou et al.\ from UK Biobank Olink proteomics and individual-level T1D genotype--phenotype data |
| `LD_Eur_18mvariants/int8/` (genoscores) | Block-LD matrix for UK Biobank participants of European ancestry; $N = 362{,}063$ samples; ${\sim}18$ million variants; GRCh37 coordinates; quantized to int8 | UK Biobank; computed using magenpy |
| `refpop/kg.2020.hg38.eur.bim` | Variant index (rsid, chromosome, position hg38) for the 1000 Genomes EUR reference panel; 503 samples; used for (chr, pos) $\to$ rsid lookup | 1000 Genomes Project phase 3, GRCh38 liftover |
| `refGene.txt.gz` | RefSeq protein-coding gene annotations (NM\_ accessions); used to label loci with the nearest gene | UCSC Genome Browser, hg38 assembly |

# Individual-level T1D outcome (Zhou et al.)

`r nrow(indiv_dt)` scalar instruments from the UK Biobank Olink proteomics study
($N = 33{,}902$), with instrument-outcome coefficients estimated from
individual-level genotype-outcome data in the T1D case-control study (Zhou et al.).
Nearest Gene lists all protein-coding genes at the locus.

\newpage

```{r instruments-table-indiv, results='asis'}
make_table(indiv_dt, label = "instruments-indiv",
           caption_extra = " --- UK Biobank, individual-level T1D outcome (Zhou et al.)",
           has_eff_rank = FALSE)
```

# UK Biobank --- summary-level T1D outcome

`r nrow(coeffs_ukb)` scalar instruments from the UK Biobank Olink proteomics
study ($N = 33{,}902$). Instrument-outcome coefficients from Iakovliev et al.

\newpage

```{r instruments-table-ukb, results='asis'}
make_table(coeffs_ukb, label = "instruments-ukb",
           caption_extra = " --- UK Biobank SNP--protein coefficients, summary-level T1D",
           has_sd_z = "sd.Z" %in% names(coeffs_ukb),
           scale_by_sdz = TRUE,
           small_font = TRUE)
```

\clearpage

```{r scatter-indiv, fig.cap=paste0("UK Biobank, individual-level T1D (Zhou et al.): $\\hat{\\gamma}$ vs $\\hat{\\alpha}$ (both per SD of scalar instrument) for ", nrow(indiv_dt), " instruments. IVW slope estimated from trans-pQTLs only: $\\hat\\theta = ", round(ivw_indiv_trans$Estimate, 3), "$ (SE $", round(ivw_indiv_trans$SE, 3), "$, $p = ", ivw_indiv_trans$pvalue.formatted, "$). Red dashed line: IVW slope (trans only). Red point: PDCD1 cis-pQTL."), fig.width = 7, fig.height = 6}
plot_iv_estimates(
    alpha_hat    = indiv_dt$alpha_hat,
    se.alpha_hat = indiv_dt$se.alpha_hat,
    gamma_hat    = indiv_dt$gamma_hat,
    se.gamma_hat = indiv_dt$se.gamma_hat,
    theta        = ivw_indiv_trans$Estimate,
    qtlname      = indiv_dt$gene,
    highlight    = indiv_dt$scoreid %in% cis_indiv
)
```

```{r scatter-ukb, fig.cap=paste0("UK Biobank (summary-level T1D): $\\hat{\\gamma}$ vs $\\hat{\\alpha}$ (both per SD of scalar instrument) for ", nrow(coeffs_ukb), " instruments. IVW slope estimated from trans-pQTLs only: $\\hat\\theta = ", round(ivw_ukb_trans$Estimate, 3), "$ (SE $", round(ivw_ukb_trans$SE, 3), "$, $p = ", ivw_ukb_trans$pvalue.formatted, "$). Red dashed line: IVW slope (trans only). Red point: PDCD1 cis-pQTL."), fig.width = 7, fig.height = 6}
sdz <- coeffs_ukb$sd.Z
plot_iv_estimates(
    alpha_hat    = coeffs_ukb$alpha_hat    * sdz,
    se.alpha_hat = coeffs_ukb$se.alpha_hat * sdz,
    gamma_hat    = coeffs_ukb$gamma_hat    * sdz,
    se.gamma_hat = coeffs_ukb$se.gamma_hat * sdz,
    theta        = ivw_ukb_trans$Estimate,
    qtlname      = sub(",.*$", "", coeffs_ukb$gene),
    highlight    = coeffs_ukb$qtlname == cis_ukb
)
```

# Comparison of $\hat\alpha$ coefficients across analyses

```{r alpha-compare, fig.cap="Instrument-exposure coefficient $\\hat{\\alpha}$ (per SD of scalar instrument) from the individual-level analysis (x-axis) versus the UK Biobank summary-level analysis (y-axis). Loci matched by chromosomal overlap. Grey dashed line: identity ($y=x$). Blue dotted line: OLS regression through the origin.", fig.width = 6, fig.height = 6}
dt_indiv <- indiv_dt[,  .(gene = gene, chr, lo = locus_start, hi = locus_end,
                           alpha_indiv = alpha_hat)]
dt_ukb   <- coeffs_ukb[, .(chr, lo = locus_start, hi = locus_end,
                             alpha_ukb = alpha_hat * sd.Z)]
setkeyv(dt_ukb,   c("chr", "lo", "hi"))
setkeyv(dt_indiv, c("chr", "lo", "hi"))
alpha_comp <- foverlaps(dt_indiv, dt_ukb, nomatch = 0L)

slope_ols <- coef(lm(alpha_ukb ~ 0 + alpha_indiv, data = alpha_comp))["alpha_indiv"]
ax_breaks <- pretty(c(0, max(alpha_comp$alpha_indiv, alpha_comp$alpha_ukb, na.rm = TRUE)))

ggplot(alpha_comp, aes(x = alpha_indiv, y = alpha_ukb)) +
    geom_hline(yintercept = 0, linewidth = 0.4) +
    geom_vline(xintercept = 0, linewidth = 0.4) +
    geom_abline(slope = 1,         intercept = 0, linetype = "dashed", colour = "grey50") +
    geom_abline(slope = slope_ols, intercept = 0, linetype = "dotted", colour = "blue") +
    geom_point() +
    geom_text_repel(aes(label = gene), size = 3, max.overlaps = Inf) +
    scale_x_continuous(breaks = ax_breaks, limits = range(ax_breaks)) +
    scale_y_continuous(breaks = ax_breaks, limits = range(ax_breaks)) +
    labs(x = expression(hat(alpha)[indiv]),
         y = expression(hat(alpha)[UKB])) +
    coord_fixed() +
    theme_bw()
```

# Comparison of $\hat\gamma$ coefficients across analyses

```{r gamma-compare, fig.cap="Instrument-outcome coefficient $\\hat{\\gamma}$ (per SD of scalar instrument, in log-odds units) from the individual-level analysis (x-axis) versus the UK Biobank summary-level analysis (y-axis). Loci matched by chromosomal overlap. Grey dashed line: identity ($y=x$). Blue dotted line: OLS regression through the origin.", fig.width = 6, fig.height = 6}
dt_indiv_g <- indiv_dt[,  .(gene = gene, chr, lo = locus_start, hi = locus_end,
                              gamma_indiv = gamma_hat)]
dt_ukb_g   <- coeffs_ukb[, .(chr, lo = locus_start, hi = locus_end,
                               gamma_ukb = gamma_hat * sd.Z)]
setkeyv(dt_ukb_g,   c("chr", "lo", "hi"))
setkeyv(dt_indiv_g, c("chr", "lo", "hi"))
gamma_comp <- foverlaps(dt_indiv_g, dt_ukb_g, nomatch = 0L)

slope_ols_g <- coef(lm(gamma_ukb ~ 0 + gamma_indiv, data = gamma_comp))["gamma_indiv"]
ax_lim_g    <- max(abs(c(gamma_comp$gamma_indiv, gamma_comp$gamma_ukb)), na.rm = TRUE) * 1.05
ax_breaks_g <- pretty(c(-ax_lim_g, ax_lim_g))

ggplot(gamma_comp, aes(x = gamma_indiv, y = gamma_ukb)) +
    geom_hline(yintercept = 0, linewidth = 0.4) +
    geom_vline(xintercept = 0, linewidth = 0.4) +
    geom_abline(slope = 1,           intercept = 0, linetype = "dashed", colour = "grey50") +
    geom_abline(slope = slope_ols_g, intercept = 0, linetype = "dotted", colour = "blue") +
    geom_point() +
    geom_text_repel(aes(label = gene), size = 3, max.overlaps = Inf) +
    scale_x_continuous(breaks = ax_breaks_g, limits = range(ax_breaks_g)) +
    scale_y_continuous(breaks = ax_breaks_g, limits = range(ax_breaks_g)) +
    labs(x = expression(hat(gamma)[indiv]),
         y = expression(hat(gamma)[UKB])) +
    coord_fixed() +
    theme_bw()
```

<!-- SCALLOP/UKBB section commented out
# SCALLOP/UKBB --- summary-level T1D outcome

```{r instruments-table-sc, eval=FALSE, include=FALSE}
make_table(coeffs_sc, label = "instruments-sc",
           caption_extra = " --- SCALLOP/UKBB SNP--protein coefficients, summary-level T1D")
```

```{r scatter-sc, eval=FALSE, include=FALSE}
plot_iv_estimates(
    alpha_hat    = coeffs_sc$alpha_hat,
    se.alpha_hat = coeffs_sc$se.alpha_hat,
    gamma_hat    = coeffs_sc$gamma_hat,
    se.gamma_hat = coeffs_sc$se.gamma_hat,
    theta        = ivw_sc$Estimate,
    qtlname      = coeffs_sc$gene,
    highlight    = coeffs_sc$qtlname == cis_sc
)
```
-->

```{r mrhevo-results, results='asis'}
if (!file.exists("mrhevo_pdcd1_t1d_fit.rds")) {
    cat("\\textit{MR-Hevo posterior results not yet available.",
        "Run \\texttt{run\\_mrhevo\\_pdcd1\\_t1d.R} to generate",
        "\\texttt{mrhevo\\_pdcd1\\_t1d\\_fit.rds}.}\n")
}
```
