---
title: "Summary-statistics Mendelian randomization and GATE score"
subtitle: "Methods and application to PDCD1 protein and type 1 diabetes"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    latex_engine: pdflatex
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{caption}
  - \captionsetup[longtable]{width=\linewidth}
  - \usepackage{url}
  - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.pos = "H")
options(knitr.kable.NA = "")
library(data.table)
library(ggplot2)
library(ggrepel)
library(knitr)
library(kableExtra)
devtools::load_all(".", quiet = TRUE)
```

\part{Theoretical derivations}

# Recovering multivariate regression coefficients from univariate summary statistics

## Introduction

In genome-wide association studies (GWAS), summary statistics are typically reported as
*marginal* (univariate) regression coefficients $\hat\beta_{u,j}$, one per SNP,
estimated by regressing the trait on each SNP separately.  When a locus contains
$P$ SNPs in linkage disequilibrium (LD), the marginal coefficient for SNP $j$ conflates
its direct effect on the trait with indirect contributions absorbed from correlated
neighbours.

This section derives the formula for recovering *multivariate* (joint) regression
coefficients $\hat{\bm\beta}_m$ from the marginal coefficients, using the LD correlation
matrix among SNPs.  The derivation works with **unstandardized** genotype units
(allele dosage 0/1/2) throughout: both the input ($\hat{\bm\beta}_u$) and the output
($\hat{\bm\beta}_m$) are in units of trait change per additional copy of the effect allele.

## Notation

Let $n$ be the number of individuals and $P$ the number of SNPs in a locus.

- $\sigma_j^2 = \operatorname{Var}(g_j) = 2f_j(1 - f_j)$: genotype variance for SNP $j$.
- $\mathbf{D} = \operatorname{diag}(\sigma_1, \dots, \sigma_P)$: diagonal matrix of SDs.
- $\mathbf{G}^* = \mathbf{G}\mathbf{D}^{-1}$: genotype matrix scaled to unit variance per column.
- $\mathbf{C}_g = \mathbf{G}^{*\top}\mathbf{G}^*/n$: $P\times P$ LD correlation matrix.
- $\boldsymbol{\Sigma}_g = \mathbf{G}^\top\mathbf{G}/n = \mathbf{D}\mathbf{C}_g\mathbf{D}$: genotype covariance matrix.

## Key identity

The marginal-to-multivariate relationship follows from the normal equations
$\boldsymbol{\Sigma}_g\hat{\bm\beta}_m = (1/n)\mathbf{G}^\top\mathbf{y} = \mathbf{D}^2\hat{\bm\beta}_u$:

$$
\boxed{
\hat{\bm\beta}_m = \mathbf{D}^{-1}\,\mathbf{C}_g^{-1}\,\mathbf{D}\,\hat{\bm\beta}_u.
}
\tag{1}
$$

Both $\hat{\bm\beta}_u$ and $\hat{\bm\beta}_m$ are in **units of trait change per allele**;
$\mathbf{D}$ converts between per-allele and per-SD units internally.
Equivalently: scale marginal coefficients to per-SD units, apply $\mathbf{C}_g^{-1}$, rescale back.

Verification: $\boldsymbol{\Sigma}_g\hat{\bm\beta}_m = \mathbf{D}\mathbf{C}_g\mathbf{D}\cdot\mathbf{D}^{-1}\mathbf{C}_g^{-1}\mathbf{D}\hat{\bm\beta}_u = \mathbf{D}^2\hat{\bm\beta}_u = (1/n)\mathbf{G}^\top\mathbf{y}$. $\checkmark$

## Application to GWAS summary statistics

In practice $\hat{\mathbf{C}}_g$ is estimated from a reference panel.  When $P \gg n_\text{ref}$
or SNPs are in near-perfect LD, the **truncated pseudoinverse** is used: retain only the
$K$ eigenvalues $d_k > \epsilon_0 d_1$ (default $\epsilon_0 = 0.01$):
$$
\hat{\bm\beta}_m = \hat{\mathbf{D}}^{-1}\,\hat{\mathbf{C}}_g^{+}\,\hat{\mathbf{D}}\,\hat{\bm\beta}_u.
\tag{2}
$$
The effective rank $K$ is reported for each locus.

## Scalar instrument construction

The $P$ per-allele multivariate coefficients are combined into a scalar instrument:
$$
Z_i = \frac{S_i}{\hat\alpha}, \qquad S_i = \mathbf{g}_i^\top\hat{\bm\beta}_m,
\qquad \hat\alpha = \|\hat{\bm\beta}_m\|.
$$
The regression coefficient of exposure $X$ on $Z$ equals $\hat\alpha$, confirming
$\mathbb{E}[X \mid Z] \approx \hat\alpha Z$.

## Standard error of $\hat\alpha$

The Fisher information from $N$ observations of $X = \alpha Z + \varepsilon_X$ is
$\mathcal{I}_\alpha = N\,\operatorname{Var}(Z)/(\sigma_X^2 - \operatorname{Var}(S))$,
where $\sigma_X^2$ is estimated from per-SNP marginal SEs via
$\sigma_X^2 \approx \sigma_j^2(N\operatorname{SE}(\hat\beta_{u,j})^2 + \hat\beta_{u,j}^2)$
(median over $j$).  The SE is

$$
\boxed{
\operatorname{SE}(\hat\alpha)
  = \sqrt{\frac{\sigma_X^2 - \widehat{\operatorname{Var}}(S)}
               {N\,\widehat{\operatorname{Var}}(Z)}}.
}
\tag{3}
$$

# Instrument-outcome coefficient $\hat\gamma$

## Multivariate outcome coefficients

The per-allele univariate SNP-outcome coefficients $\hat{\bm\gamma}_u$ are recovered from
$z$-scores as $\hat\gamma_{u,j} = z_j / \sqrt{N_Y\sigma_j^2 p(1-p)}$.
The same LD inversion gives the multivariate outcome coefficients:
$$
\hat{\bm\gamma}_m = \hat{\mathbf{D}}^{-1}\,\hat{\mathbf{C}}_g^{+}\,\hat{\mathbf{D}}\,\hat{\bm\gamma}_u,
\tag{4}
$$
using the same $\hat{\mathbf{D}}$, $\hat{\mathbf{C}}_g^+$ as for the exposure.

## Closed-form estimator

Projecting the outcome score $T_i = \mathbf{g}_i^\top\hat{\bm\gamma}_m$ onto the scalar
instrument $Z$ gives
$$
\hat\gamma
  = \hat\alpha\,\frac{\hat{\bm\gamma}_m^\top\hat{\boldsymbol\Sigma}_g\hat{\bm\beta}_m}
                     {\hat{\bm\beta}_m^\top\hat{\boldsymbol\Sigma}_g\hat{\bm\beta}_m}
  = \hat\alpha\,
    \frac{\displaystyle\sum_{j=1}^P \hat\sigma_j^2\,\hat\gamma_{u,j}\,\hat\beta_{m,j}}
         {\widehat{\operatorname{Var}}(S)}.
\tag{5}
$$
$\hat\gamma$ is a weighted sum of the per-allele products $\hat\gamma_{u,j}\hat\beta_{m,j}$
with weights proportional to genotype variance $\hat\sigma_j^2$.

## Standard error of $\hat\gamma$

For a binary outcome (proportion of cases $p$):
$$
\boxed{
\operatorname{SE}(\hat\gamma)
  = \frac{1}{\sqrt{N_Y\,\widehat{\operatorname{Var}}(Z)\,p(1-p)}}
  = \frac{\hat\alpha}{\sqrt{N_Y\,\widehat{\operatorname{Var}}(S)\,p(1-p)}}.
}
\tag{6}
$$
Here $\widehat{\operatorname{Var}}(Z) = \widehat{\operatorname{Var}}(S)/\hat\alpha^2$ is
computed from the reference panel.

# GATE score coefficient from summary-level pQTL statistics

## The GATE score

The GATE (Genomewide Aggregated Trans- Effects) score sums genetically predicted
exposure values over all trans-pQTLs:
$$
G_i = \sum_{k \in \text{trans}} \hat\alpha_k\, Z_{ik},
\qquad
G_i^* = G_i \,/\, \mathrm{SD}_\mathrm{ref}(G),
$$
where $\mathrm{SD}_\mathrm{ref}(G) = \sqrt{\sum_k \hat\alpha_k^2\, \mathrm{sd}_{Z,k}^2}$
(trans-loci are independent so cross-terms vanish).

## Score test for $\hat\beta_\mathrm{GATE}$

Under the null logistic model with residuals $r_i = Y_i - \hat\mu_i^{(0)}$ and
weights $w_i = \hat\mu_i^{(0)}(1-\hat\mu_i^{(0)})$, the composite score and
partial information are (using independence across loci):
$$
U_{G^*} = \frac{1}{\mathrm{SD}_G} \sum_k \hat\alpha_k\, U_k, \qquad
V_{G^*} = \frac{1}{\mathrm{SD}_G^2} \sum_k \hat\alpha_k^2\, V_k.
$$
Substituting $U_k = \hat\gamma_k V_k$ (score estimate for locus $k$) and
$V_k = 1/\mathrm{SE}(\hat\gamma_k)^2$:

$$
\boxed{
\hat\beta_\mathrm{GATE}
= \mathrm{SD}_G \cdot
  \frac{\displaystyle\sum_k \hat\alpha_k\, \hat\gamma_k\, /\, \mathrm{SE}(\hat\gamma_k)^2}
       {\displaystyle\sum_k \hat\alpha_k^2\, /\, \mathrm{SE}(\hat\gamma_k)^2},
\qquad
\mathrm{SE}(\hat\beta_\mathrm{GATE})
= \frac{\mathrm{SD}_G}
       {\sqrt{\displaystyle\sum_k \hat\alpha_k^2\, /\, \mathrm{SE}(\hat\gamma_k)^2}}.
}
\tag{7}
$$

## Simplification of the SE

Substituting $V_k = N\,\mathrm{sd}_{Z,k}^2\,p(1-p)$ from eq.\ (6) gives
$\sum_k \hat\alpha_k^2 V_k = N p(1-p)\,\mathrm{SD}_G^2$, so:
$$
\mathrm{SE}(\hat\beta_\mathrm{GATE}) = \frac{1}{\sqrt{N\,p(1-p)}}.
\tag{8}
$$
The SE equals that of a single 1-df binary logistic regression.
The z-statistic simplifies to the IVW z-statistic with weights
$w_k = (\hat\alpha_k^*)^2$ (where $\hat\alpha_k^* = \hat\alpha_k\,\mathrm{sd}_{Z,k}$
are the per-SD(Z) exposure coefficients):
$$
z_\mathrm{GATE}
= \sqrt{N\,p(1-p)}\cdot
  \frac{\sum_k (\hat\alpha_k^*)^2\, \hat\theta_k}{\sqrt{\sum_k (\hat\alpha_k^*)^2}},
\qquad \hat\theta_k = \hat\gamma_k^*/\hat\alpha_k^*.
\tag{9}
$$

\cleardoublepage

\part{PDCD1/T1D MR-Hevo analysis}

```{r load-data}
## ---- Load protein-coding gene spans (used for NA gene filling) ----
genes_coding <- fread("refGene.txt.gz", header = FALSE,
                      select = c(2L, 3L, 5L, 6L, 13L),
                      col.names = c("accession", "chrom", "txStart", "txEnd", "symbol"))
genes_coding[, chr_g := suppressWarnings(as.integer(sub("^chr", "", chrom)))]
genes_coding <- genes_coding[!is.na(chr_g) & startsWith(accession, "NM_")]
genes_coding <- genes_coding[, .(txStart = min(txStart), txEnd = max(txEnd)),
                               by = .(chr = chr_g, symbol)]

fill_na_genes <- function(dt) {
    na_idx <- which(is.na(dt$gene))
    if (length(na_idx) == 0L) return(dt)
    for (i in na_idx) {
        chr_i   <- as.integer(dt$chr[i])
        start_i <- dt$locus_start[i]
        end_i   <- dt$locus_end[i]
        hits <- genes_coding[chr == chr_i &
                              ((txStart >= start_i & txStart <= end_i) |
                               (txEnd   >= start_i & txEnd   <= end_i))]
        if (nrow(hits) > 0L)
            dt$gene[i] <- paste(unique(hits$symbol), collapse = ", ")
    }
    dt
}

find_cis <- function(dt) {
    pdcd1_mid <- 241858000L
    chr2 <- dt[chr == "2"]
    if (nrow(chr2) == 0L) return(character(0))
    chr2$qtlname[which.min(abs(chr2$locus_start - pdcd1_mid))]
}

make_coeffs <- function(alpha_rds, gamma_rds, genes_rds) {
    alpha_dt <- readRDS(alpha_rds)
    gamma_dt  <- readRDS(gamma_rds)
    extra_cols <- intersect(c("sd.Z", "n_ld_blocks", "min_maf"), names(alpha_dt))
    alpha_sel  <- c("qtlname", "chr", "locus_start", "locus_end",
                    "n_snps", "eff_rank", "alpha_hat", "se.alpha_hat", extra_cols)
    dt <- merge(
        alpha_dt[, ..alpha_sel],
        gamma_dt[, .(qtlname, gamma_hat, se.gamma_hat)],
        by = "qtlname"
    )
    dt[, chr := as.character(chr)]
    dt <- get_coeffratios(dt, use.delta = TRUE)
    genes <- readRDS(genes_rds)
    dt <- merge(dt, genes, by = "qtlname", all.x = TRUE)
    dt <- fill_na_genes(dt)
    dt[, chr_sort := suppressWarnings(as.integer(chr))]
    dt[chr == "X", chr_sort := 23L]
    dt[chr == "Y", chr_sort := 24L]
    setorder(dt, chr_sort, locus_start)
    dt[, chr_sort := NULL]
    dt
}

coeffs_ukb <- make_coeffs("alpha_pdcd1_UKB.rds",
                            "gamma_t1d_ukb.rds",
                            "locus_genes_ukb.rds")
cis_ukb       <- find_cis(coeffs_ukb)
ivw_ukb       <- get_estimatorsMR(coeffs_ukb)
ivw_ukb_trans <- get_estimatorsMR(coeffs_ukb[qtlname != cis_ukb])

indiv_raw <- fread("pdcd1.mr.coeffs.csv")
setnames(indiv_raw,
         c("chrom", "startpos", "endpos", "alpha.hat", "se.alpha.hat"),
         c("chr",   "locus_start", "locus_end", "alpha_hat", "se.alpha_hat"))
indiv_raw[, chr := as.character(chr)]
indiv_raw[, gene := sub(",.*$", "", qtlname)]
indiv_raw[, chr_sort := suppressWarnings(as.integer(chr))]
indiv_raw[chr == "X", chr_sort := 23L]
indiv_raw[chr == "Y", chr_sort := 24L]
setorder(indiv_raw, chr_sort, locus_start)
indiv_raw[, chr_sort := NULL]
indiv_dt <- get_coeffratios(indiv_raw, use.delta = TRUE)

gene_overrides <- list(
    list(chr =  1L, lo = 113e6, hi = 115e6, gene = "PTPN22"),
    list(chr =  2L, lo = 240e6, hi = 243e6, gene = "PDCD1"),
    list(chr =  9L, lo =   5e6, hi =   6e6, gene = "CD274"),
    list(chr = 10L, lo =   5e6, hi =   7e6, gene = "IL15RA"),
    list(chr = 12L, lo = 109e6, hi = 113e6, gene = "SH2B3"),
    list(chr = 19L, lo =  48e6, hi =  50e6, gene = "FUT2")
)
apply_gene_overrides <- function(dt) {
    for (ov in gene_overrides) {
        idx <- which(as.integer(dt$chr) == ov$chr &
                     dt$locus_start >= ov$lo &
                     dt$locus_start <= ov$hi)
        if (length(idx)) set(dt, idx, "gene", ov$gene)
    }
    invisible(dt)
}
apply_gene_overrides(coeffs_ukb)
apply_gene_overrides(indiv_dt)

cis_indiv       <- indiv_dt[qtl_type == "cis-x", scoreid]
ivw_indiv       <- get_estimatorsMR(indiv_dt)
ivw_indiv_trans <- get_estimatorsMR(indiv_dt[!(scoreid %in% cis_indiv)])

comp_direct <- readRDS("gamma_comparison_analytic_vs_direct.rds")
r_ad <- cor(comp_direct$gamma_analytic_sdz, comp_direct$gamma_direct,
            use = "complete.obs")
comp_direct <- merge(comp_direct,
                     coeffs_ukb[, .(qtlname, chr, locus_start, locus_end)],
                     by.x = "locus_id", by.y = "qtlname", all.x = TRUE)
comp_direct[, chr := as.character(chr)]

coeffs_gs     <- make_coeffs("alpha_pdcd1_UKB.rds",
                              "gamma_t1d_gs.rds",
                              "locus_genes_gs.rds")
cis_gs        <- find_cis(coeffs_gs)
ivw_gs        <- get_estimatorsMR(coeffs_gs)
ivw_gs_trans  <- get_estimatorsMR(coeffs_gs[qtlname != cis_gs])
apply_gene_overrides(coeffs_gs)
```

```{r instruments-chunk, echo=FALSE}
make_table <- function(dt, label, caption_extra = "",
                        has_eff_rank = TRUE, has_sd_z = FALSE,
                        scale_by_sdz = FALSE, small_font = FALSE,
                        has_ld_blocks = FALSE) {
    sdz <- if (scale_by_sdz && "sd.Z" %in% names(dt)) dt$sd.Z else 1
    base_cols <- list(
        `Nearest Gene`         = dt$gene,
        Chr                    = dt$chr,
        `Pos (Mb)`             = round(dt$locus_start / 1e6, 2),
        `Width (Mb)`           = round((dt$locus_end - dt$locus_start) / 1e6, 2),
        `Num SNPs`             = dt$n_snps
    )
    if (has_eff_rank)
        base_cols[["eff rank"]] <- dt$eff_rank
    if (has_sd_z)
        base_cols[["SD($Z$)"]] <- round(dt$sd.Z, 3)
    if (has_ld_blocks)
        base_cols[["LD blks"]] <- dt$n_ld_blocks
    if ("min_maf" %in% names(dt))
        base_cols[["Min MAF"]] <- round(dt$min_maf, 2)
    base_cols[["$\\hat\\alpha$ (SE)"]] <- sprintf("%.4f (%.4f)", dt$alpha_hat * sdz, dt$se.alpha_hat * sdz)
    base_cols[["$\\hat\\gamma$ (SE)"]] <- sprintf("%.4f (%.4f)", dt$gamma_hat * sdz, dt$se.gamma_hat * sdz)
    base_cols[["*"]]                   <- ifelse(abs(dt$gamma_hat / dt$se.gamma_hat) > qnorm(0.995), "*", "")
    tab <- as.data.table(base_cols)

    eff_rank_note  <- if (has_eff_rank)             "eff rank: effective rank of LD matrix. " else ""
    sd_z_note      <- if (has_sd_z)                 "SD($Z$): standard deviation of scalar instrument in reference panel. " else ""
    ld_blocks_note <- if (has_ld_blocks)             "LD blks: number of zarr LD blocks spanned by the locus. " else ""
    min_maf_note   <- if ("min_maf" %in% names(dt)) "Min MAF: minimum minor allele frequency across SNPs contributing to the locus (UKBB proteomics GWAS). " else ""
    k <- kable(tab, format = "latex", booktabs = TRUE, escape = FALSE,
          longtable = TRUE,
          caption = paste0(
              "Instrument coefficients", caption_extra, ". ",
              "Nearest Gene: protein-coding gene(s) nearest to or overlapping the locus. ",
              "Num SNPs: SNPs in locus; ", eff_rank_note, sd_z_note, ld_blocks_note, min_maf_note,
              "$\\hat\\alpha$ and $\\hat\\gamma$: instrument-exposure and ",
              "instrument-outcome coefficients per SD of the scalar instrument (SE in parentheses). ",
              "*: $|\\hat\\gamma / \\mathrm{SE}(\\hat\\gamma)| > 2.58$ ",
              "(two-sided $p < 0.01$)."
          ),
          label = label) |>
        kable_styling(latex_options = c("repeat_header", "striped"),
                      font_size = if (small_font) 8 else NULL) |>
        column_spec(1, width = "2.5cm") |>
        column_spec(3:5, width = "1.1cm")
    col_n <- 6L
    if (has_eff_rank) { k <- k |> column_spec(col_n, width = "1.1cm"); col_n <- col_n + 1L }
    if (has_sd_z)      col_n <- col_n + 1L
    if (has_ld_blocks) k <- k |> column_spec(col_n, width = "1.1cm")
    k
}
```

# Introduction

We estimate the causal effect of circulating PDCD1 (programmed cell death protein 1)
levels on risk of type 1 diabetes (T1D) using Mendelian randomization with the
regularized horseshoe prior (MR-Hevo).

The T1D outcome dataset is the same across all three analyses: cases from the
Scottish T1D Bioresource (SDRNT1BIO) and controls from Generation Scotland.

Results are presented from three analyses, differing in the LD reference panel
used to construct scalar instruments and in whether instrument-outcome
coefficients $\hat\gamma$ are estimated from individual-level genotypes or
from GWAS summary statistics:

1. **Analysis 1 (Zhou et al.)**: $\hat\alpha$ and $\hat\gamma$ from Zhou et al.,
   who used the 1000 Genomes EUR panel (503 samples) as the LD reference.

2. **Analysis 2 (directly computed $\hat\gamma$, UKBB 18M EUR LD)**: $\hat\alpha$
   from UK Biobank Olink proteomics ($N = 33{,}902$); $\hat\gamma$ computed by
   scoring each individual on $Z_i = \sum_j G_{ij}\,\hat\beta_{m,j}$ and regressing
   T1D on $Z_i$ (4,922 cases, 7,452 controls; adjusted for sex, age, and the first
   three principal components). Instrument weights $\hat\beta_m$ from the UKBB EUR
   LD matrix (362,063 samples, 18 million variants).

3. **Analysis 3 (LD approximate $\hat\gamma$, UKBB 18M EUR LD)**: same $\hat\alpha$
   and instrument weights $\hat\beta_m$ as Analysis 2; $\hat\gamma$ estimated from
   per-SNP SNP-outcome coefficients via the LD approximate projection formula
   (Section~\ref{instrument-outcome-coefficient-hatgamma}), using the same UKBB EUR LD matrix.

Scalar genetic instruments are constructed from loci associated with PDCD1 at
$p < 10^{-6}$, excluding the HLA region (chromosome 6: 25--34 Mb), with minor
allele frequency $\geq 0.01$.
All displayed values of $\hat\alpha$ and $\hat\gamma$ are in per-SD($Z$) units.

# Data sources

```{r datasources-table, results='asis', echo=FALSE}
ds <- data.frame(
  File = c(
    "\\path{PDCD1_Q15116_OID21396_v1_Oncology.tar}",
    "\\path{z\\_scores\\_gs.rds}",
    "\\path{pdcd1.mr.coeffs.csv}",
    "\\path{LD_Eur_18mvariants/int8/} (genoscores)",
    "\\path{GS_T1biochrall_filtered_rsmapped.bed/.bim/.fam} (diabepi)",
    "\\path{PHENO_T1DM\\_...NORELPCA.samples} (diabepi)",
    "\\path{refpop/kg.2020.hg38.eur.bim}",
    "\\path{refGene.txt.gz}"
  ),
  Description = c(
    "SNP--protein GWAS summary statistics for PDCD1 (UniProt Q15116; Olink Oncology panel, assay OID21396); $N = 33{,}902$ UK Biobank EUR participants",
    "Per-SNP T1D SNP-outcome z-statistics computed from individual-level SDRNT1BIO/GS genotypes by score test; 4,922 cases, 7,452 controls; aligned to \\texttt{bim\\_a1} allele convention",
    "Scalar $\\hat\\alpha$ and $\\hat\\gamma$ coefficients from Analysis 1 (Zhou et al.), with 1000 Genomes EUR LD",
    "Block-LD matrix; UKBB EUR participants; $N = 362{,}063$; $\\sim$18M variants; GRCh37; int8",
    "Individual-level T1D genotypes (SDRNT1BIO cases + GS controls); 25,744 individuals; 7.8M SNPs; GRCh37; plink BED format",
    "T1D case-control phenotype file; 4,922 cases, 7,452 controls; includes sex, age, PC1--3 covariates; Oxford .samples format",
    "Variant index (rsid, chr, pos hg38) for 1000 Genomes EUR panel; 503 samples; used for (chr, pos) $\\to$ rsid lookup",
    "RefSeq protein-coding gene annotations (NM\\_ accessions); used to label loci with nearest gene"
  ),
  Source = c(
    "UK Biobank Pharma Proteomics Project (UKB-PPP); Sun et al.\\ (2023) \\textit{Nature} 622:329",
    "Derived from GS plink files; output of \\texttt{compute\\_gamma\\_indiv.R}",
    "Computed by Zhou et al.\\ from UKB-PPP Olink data and SDRNT1BIO/GS individual-level data",
    "UK Biobank; computed using magenpy",
    "Scottish Diabetes Research Network Type 1 Bioresource (SDRNT1BIO) and Generation Scotland (GS)",
    "SDRNT1BIO/GS; quality-controlled for relatedness, non-T1D diabetes, and population structure",
    "1000 Genomes Project phase 3, GRCh38 liftover",
    "UCSC Genome Browser, hg38 assembly"
  ),
  stringsAsFactors = FALSE
)
kable(ds, format = "latex", booktabs = TRUE, longtable = TRUE, escape = FALSE,
      caption = "Source data files used in this analysis.",
      label  = "datasources",
      col.names = c("File", "Description", "Source")) |>
  kable_styling(latex_options = c("repeat_header", "striped"), font_size = 8) |>
  column_spec(1, width = "3.5cm") |>
  column_spec(2, width = "6.5cm") |>
  column_spec(3, width = "4.5cm")
```

# Analysis 1 (Zhou et al.): individual-level $\hat\gamma$, 1000 Genomes EUR LD

`r nrow(indiv_dt)` scalar instruments; $\hat\alpha$ and $\hat\gamma$ from Zhou et al.,
computed from individual-level UK Biobank Olink proteomics data and SDRNT1BIO/GS
T1D genotype--phenotype data, using the 1000 Genomes EUR panel for LD.

\newpage

```{r instruments-table-indiv, results='asis'}
make_table(indiv_dt, label = "instruments-indiv",
           caption_extra = " --- Analysis 1 (Zhou et al.): 1000 Genomes EUR LD",
           has_eff_rank = FALSE)
```

\clearpage

```{r scatter-indiv, fig.cap=paste0("Analysis 1 (Zhou et al.), 1000 Genomes EUR LD. $\\hat{\\gamma}$ vs $\\hat{\\alpha}$ (per SD of scalar instrument) for ", nrow(indiv_dt), " instruments. IVW slope estimated from trans-pQTLs only: $\\hat\\theta = ", round(ivw_indiv_trans$Estimate, 3), "$ (SE $", round(ivw_indiv_trans$SE, 3), "$, $p = ", ivw_indiv_trans$pvalue.formatted, "$). Red dashed line: IVW slope (trans only). Red point: PDCD1 cis-pQTL."), fig.width = 7, fig.height = 6}
plot_iv_estimates(
    alpha_hat    = indiv_dt$alpha_hat,
    se.alpha_hat = indiv_dt$se.alpha_hat,
    gamma_hat    = indiv_dt$gamma_hat,
    se.gamma_hat = indiv_dt$se.gamma_hat,
    theta        = ivw_indiv_trans$Estimate,
    qtlname      = indiv_dt$gene,
    highlight    = indiv_dt$scoreid %in% cis_indiv
)
```

# Analysis 2: individual-level $\hat\gamma$ from SDRNT1BIO/GS genotypes, UKBB 18M EUR LD

`r nrow(coeffs_gs)` scalar instruments; $\hat\alpha$ from UK Biobank Olink
proteomics ($N = 33{,}902$). $\hat\gamma$ estimated from individual-level
SDRNT1BIO/GS genotypes (4,922 cases, 7,452 controls), adjusted for sex, age, and
the first three genetic principal components.  Instrument weights $\hat\beta_m$
derived from the UKBB EUR LD matrix (362,063 samples, 18 million variants).

```{r scatter-gs, fig.cap=paste0("Analysis 2: directly computed $\\hat{\\gamma}$ (individual-level logistic regression, UKBB 18M EUR LD instrument weights) vs $\\hat{\\alpha}$ (per SD of scalar instrument). IVW slope estimated from trans-pQTLs only: $\\hat\\theta = ", round(ivw_direct_trans$Estimate, 3), "$ (SE $", round(ivw_direct_trans$SE, 3), "$, $p = ", ivw_direct_trans$pvalue.formatted, "$). Red dashed line: IVW slope (trans only). Red point: PDCD1 cis-pQTL."), fig.width = 7, fig.height = 6}
d_sc <- merge(
    comp_direct[, .(locus_id, gene, gamma_direct, se_gamma_direct)],
    coeffs_ukb[, .(qtlname, alpha_hat, se.alpha_hat, sd.Z)],
    by.x = "locus_id", by.y = "qtlname"
)
d_sc[, theta_IV    := gamma_direct / (alpha_hat * sd.Z)]
d_sc[, se.theta_IV := se_gamma_direct / (alpha_hat * sd.Z)]
d_sc[, inv.var     := se.theta_IV^-2]
is_cis_direct  <- d_sc$locus_id == cis_gs
ivw_direct_all   <- d_sc[, .(Estimate = sum(theta_IV * inv.var) / sum(inv.var),
                               SE       = sqrt(1 / sum(inv.var)))]
ivw_direct_trans <- d_sc[!is_cis_direct,
                          .(Estimate = sum(theta_IV * inv.var) / sum(inv.var),
                            SE       = sqrt(1 / sum(inv.var)))]
ivw_direct_trans[, z := Estimate / SE]
ivw_direct_trans[, pvalue.formatted := format.z.aspvalue(z)]

plot_iv_estimates(
    alpha_hat    = d_sc$alpha_hat    * d_sc$sd.Z,
    se.alpha_hat = d_sc$se.alpha_hat * d_sc$sd.Z,
    gamma_hat    = d_sc$gamma_direct,
    se.gamma_hat = d_sc$se_gamma_direct,
    theta        = ivw_direct_trans$Estimate,
    qtlname      = sub(",.*$", "", d_sc$gene),
    highlight    = is_cis_direct
)
```

# Analysis 3: LD approximate $\hat\gamma$ from SNP-outcome coefficients, UKBB 18M EUR LD

`r nrow(coeffs_ukb)` scalar instruments; $\hat\alpha$ from UK Biobank Olink
proteomics ($N = 33{,}902$). $\hat\gamma$ estimated from per-SNP T1D SNP-outcome
coefficients (from SDRNT1BIO/GS individual-level logistic regression, 4,922 cases,
7,452 controls), LD-adjusted using the UKBB EUR LD matrix (362,063 samples, 18
million variants).  Same instrument weights $\hat\beta_m$ as Analysis 2.

\newpage

```{r instruments-table-ukb, results='asis'}
make_table(coeffs_ukb, label = "instruments-ukb",
           caption_extra = " --- Analysis 3: LD approximate SNP-outcome coefficients, UKBB 18M EUR LD",
           has_sd_z    = "sd.Z" %in% names(coeffs_ukb),
           scale_by_sdz  = TRUE,
           small_font    = TRUE,
           has_ld_blocks = "n_ld_blocks" %in% names(coeffs_ukb))
```

\clearpage

```{r scatter-ukb, fig.cap=paste0("Analysis 3: LD approximate $\\hat{\\gamma}$ (SNP-outcome coefficients + UKBB 18M EUR LD projection formula) vs $\\hat{\\alpha}$ (per SD of scalar instrument) for ", nrow(coeffs_ukb), " instruments. IVW slope estimated from trans-pQTLs only: $\\hat\\theta = ", round(ivw_ukb_trans$Estimate, 3), "$ (SE $", round(ivw_ukb_trans$SE, 3), "$, $p = ", ivw_ukb_trans$pvalue.formatted, "$). Red dashed line: IVW slope (trans only). Red point: PDCD1 cis-pQTL."), fig.width = 7, fig.height = 6}
sdz <- coeffs_ukb$sd.Z
plot_iv_estimates(
    alpha_hat    = coeffs_ukb$alpha_hat    * sdz,
    se.alpha_hat = coeffs_ukb$se.alpha_hat * sdz,
    gamma_hat    = coeffs_ukb$gamma_hat    * sdz,
    se.gamma_hat = coeffs_ukb$se.gamma_hat * sdz,
    theta        = ivw_ukb_trans$Estimate,
    qtlname      = sub(",.*$", "", coeffs_ukb$gene),
    highlight    = coeffs_ukb$qtlname == cis_ukb
)
```

# Comparison of $\hat\alpha$: Analysis 1 (1000G LD) vs Analyses 2 and 3 (UKBB 18M LD)

```{r alpha-compare, fig.cap="Instrument-exposure coefficient $\\hat{\\alpha}$ (per SD of scalar instrument): Analysis 1 / Zhou et al.\\ with 1000 Genomes EUR LD (x-axis) vs Analysis 2 with UKBB 18M EUR LD (y-axis). Loci matched by chromosomal overlap. Labelled points are outliers from the identity line ($|y-x| > 2\\,\\mathrm{SD}$). Grey dashed line: identity ($y=x$). Blue dotted line: OLS regression through the origin.", fig.width = 6, fig.height = 6}
dt_indiv <- indiv_dt[,  .(gene = gene, chr, lo = locus_start, hi = locus_end,
                           alpha_indiv = alpha_hat)]
dt_ukb   <- coeffs_ukb[, .(chr, lo = locus_start, hi = locus_end,
                             alpha_ukb   = alpha_hat * sd.Z,
                             n_ld_blocks = n_ld_blocks)]
setkeyv(dt_ukb,   c("chr", "lo", "hi"))
setkeyv(dt_indiv, c("chr", "lo", "hi"))
alpha_comp <- foverlaps(dt_indiv, dt_ukb, nomatch = 0L)

slope_ols <- coef(lm(alpha_ukb ~ 0 + alpha_indiv, data = alpha_comp))["alpha_indiv"]
ax_breaks <- pretty(c(0, max(alpha_comp$alpha_indiv, alpha_comp$alpha_ukb, na.rm = TRUE)))

alpha_comp[, resid_id := alpha_ukb - alpha_indiv]
alpha_comp[, outlier_label := ifelse(
    abs(resid_id) > 2 * sd(resid_id, na.rm = TRUE), gene, "")]

ggplot(alpha_comp, aes(x = alpha_indiv, y = alpha_ukb)) +
    geom_hline(yintercept = 0, linewidth = 0.4) +
    geom_vline(xintercept = 0, linewidth = 0.4) +
    geom_abline(slope = 1,         intercept = 0, linetype = "dashed", colour = "grey50") +
    geom_abline(slope = slope_ols, intercept = 0, linetype = "dotted", colour = "blue") +
    geom_point() +
    geom_text_repel(aes(label = outlier_label), size = 3, max.overlaps = Inf) +
    scale_x_continuous(breaks = ax_breaks, limits = range(ax_breaks)) +
    scale_y_continuous(breaks = ax_breaks, limits = range(ax_breaks)) +
    labs(x = expression(hat(alpha)[indiv]),
         y = expression(hat(alpha)[UKB])) +
    coord_fixed() +
    theme_bw()
```

# Comparison of directly computed $\hat\gamma$: Zhou et al.\ (1000G LD) vs Analysis 2 (UKBB 18M LD)

Both analyses use individual-level SDRNT1BIO/GS genotypes to compute the scalar
instrument score $Z_i = \sum_j G_{ij}\,\hat\beta_{m,j}$ for each individual and
regress T1D on $Z_i$.  They differ only in the LD reference panel used to derive
the instrument weights $\hat\beta_m$: 1000 Genomes EUR (503 samples, Analysis 1)
vs UKBB EUR 18M variant LD matrix (362,063 samples, Analysis 2).

```{r gamma-direct-direct, fig.cap="Both axes show directly computed $\\hat{\\gamma}$ from individual-level logistic regression of T1D on the scalar instrument score (per SD, log-odds). x-axis: Zhou et al.\\ Analysis 1 using 1000 Genomes EUR LD weights. y-axis: new analysis using UKBB 18M EUR LD weights (Analysis 2). Both analyses use the same SDRNT1BIO/GS individual-level genotypes; they differ only in the LD reference panel used to derive instrument weights $\\hat\\beta_m$. Loci matched by chromosomal overlap. The six largest outliers from the identity line are labelled. Grey dashed line: identity ($y=x$).", fig.width = 6, fig.height = 6}
dt_zhou <- indiv_dt[, .(gene = gene, chr, lo = locus_start, hi = locus_end,
                         gamma_zhou = gamma_hat)]
dt_new  <- comp_direct[!is.na(locus_start),
                        .(chr, lo = locus_start, hi = locus_end,
                          gamma_new_direct = gamma_direct)]
setkeyv(dt_new,  c("chr", "lo", "hi"))
setkeyv(dt_zhou, c("chr", "lo", "hi"))
dd_comp <- foverlaps(dt_zhou, dt_new, nomatch = 0L)

ax_brk_dd <- pretty(range(c(dd_comp$gamma_new_direct, dd_comp$gamma_zhou), na.rm = TRUE))

dd_comp[, resid_id := gamma_new_direct - gamma_zhou]
top6_genes <- dd_comp[order(-abs(resid_id))][seq_len(min(6L, .N)), gene]
dd_comp[, outlier_label := ifelse(gene %in% top6_genes, gene, "")]

ggplot(dd_comp, aes(x = gamma_zhou, y = gamma_new_direct)) +
    geom_hline(yintercept = 0, linewidth = 0.4) +
    geom_vline(xintercept = 0, linewidth = 0.4) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
    geom_point() +
    geom_text_repel(aes(label = outlier_label), size = 3, max.overlaps = Inf) +
    scale_x_continuous(breaks = ax_brk_dd, limits = range(ax_brk_dd)) +
    scale_y_continuous(breaks = ax_brk_dd, limits = range(ax_brk_dd)) +
    labs(x = expression(hat(gamma)["Zhou et al., 1000G LD"]),
         y = expression(hat(gamma)["direct, UKBB 18M LD"])) +
    coord_fixed() +
    theme_bw()
```

# Comparison of $\hat\gamma$: direct individual-level regression vs LD approximate

The direct method (Analysis 2, x-axis, ground truth) regresses T1D on individual
instrument scores using UKBB EUR LD weights.
The LD approximate method (Analysis 3, y-axis) estimates $\hat\gamma$ from the
same per-SNP SNP-outcome coefficients via the projection formula in
Section~\ref{instrument-outcome-coefficient-hatgamma}.
Both use identical instrument weights $\hat\beta_m$.
Attenuation of the LD approximate estimate reflects mismatch between the UKBB
reference panel LD and the true LD in the GS target dataset.
The Pearson correlation is $r = `r round(r_ad, 3)`$.

```{r gamma-analytic-vs-direct, fig.cap=paste0("Direct individual-level regression $\\hat{\\gamma}$ (x-axis, ground truth) vs LD approximate $\\hat{\\gamma}$ (y-axis). Both per SD of scalar instrument, log-odds units; same UKBB 18M EUR LD weights used for both. Pearson $r = ", round(r_ad, 3), "$. The six largest outliers from the identity line are labelled. Grey dashed: identity ($y = x$)."), fig.width = 6, fig.height = 6}
comp_direct[, resid_id := gamma_analytic_sdz - gamma_direct]
top6_genes_ad <- comp_direct[order(-abs(resid_id))][seq_len(min(6L, .N)), gene]
comp_direct[, outlier_label := ifelse(gene %in% top6_genes_ad, gene, "")]

ax_brk_ad <- pretty(range(c(comp_direct$gamma_direct,
                             comp_direct$gamma_analytic_sdz), na.rm = TRUE))

ggplot(comp_direct, aes(x = gamma_direct, y = gamma_analytic_sdz)) +
    geom_hline(yintercept = 0, linewidth = 0.4) +
    geom_vline(xintercept = 0, linewidth = 0.4) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
    geom_point() +
    geom_text_repel(aes(label = outlier_label), size = 3, max.overlaps = Inf) +
    scale_x_continuous(breaks = ax_brk_ad, limits = range(ax_brk_ad)) +
    scale_y_continuous(breaks = ax_brk_ad, limits = range(ax_brk_ad)) +
    labs(x = expression(hat(gamma)["direct (individual-level)"]),
         y = expression(hat(gamma)["LD approximate"])) +
    coord_fixed() +
    theme_bw()
```

```{r mrhevo-results, results='asis'}
if (!file.exists("mrhevo_pdcd1_t1d_fit.rds")) {
    cat("\\textit{MR-Hevo posterior results not yet available.",
        "Run \\texttt{run\\_mrhevo\\_pdcd1\\_t1d.R} to generate",
        "\\texttt{mrhevo\\_pdcd1\\_t1d\\_fit.rds}.}\n")
}
```

# GATE score: summary-level vs direct regression (PDCD1/T1D)

The summary-level GATE coefficient (eq.\ 7) is compared with the directly computed
coefficient from logistic regression of T1D on the individual-level GATE score in
the SDRNT1BIO/GS dataset (4,922 cases, 7,452 controls).
For the direct computation, the GATE score is accumulated as
$G_i = \sum_{k,j} G_{ij} \cdot (\hat\alpha_k \hat\beta_{m,kj})$ across all
trans-pQTL SNPs in the individual-level genotype data.

The two methods differ in the SD used for standardisation: the summary-level
formula uses $\mathrm{SD}_\mathrm{ref}(G)$ from the UKBB reference panel,
while the direct regression standardises by the empirical SD in the GS target
dataset.  The z-statistics are invariant to this scaling; the ratio
$\hat\beta_\mathrm{summary}/\hat\beta_\mathrm{direct}$ estimates
$\mathrm{SD}_\mathrm{ref}(G)/\mathrm{SD}_\mathrm{target}(G)$.

```{r gate-comparison-table, results='asis'}
if (file.exists("gate_results.rds")) {
    res <- readRDS("gate_results.rds")
    tab <- data.frame(
        Method          = c("Summary-level", "Direct (individual-level)"),
        `$\\hat\\beta$` = sprintf("%.4f", res$beta_GATE),
        SE              = sprintf("%.5f", res$se_GATE),
        z               = sprintf("%.3f",  res$z_GATE),
        p               = formatC(res$p_value, format="g", digits=3),
        `Trans pQTLs`   = as.integer(res$n_trans_pqtl),
        check.names = FALSE
    )
    kable(tab, format="latex", booktabs=TRUE, escape=FALSE,
          caption=paste0(
              "GATE coefficient for T1D ($\\hat\\beta_\\mathrm{GATE}$ = log-OR per SD of GATE score) ",
              "from summary-level formula (eq.\\ 7) vs direct individual-level regression. ",
              "Trans-pQTLs: loci contributing to the GATE score (PDCD1 cis-pQTL excluded). ",
              "SE (summary-level) $= 1/\\sqrt{Np(1-p)} \\approx 0.0184$ (eq.\\ 8)."),
          label="gate-comparison") |>
        kable_styling(latex_options="hold_position")
} else {
    cat("\\textit{Run \\texttt{compute\\_gate.R} to generate \\texttt{gate\\_results.rds}.}\n")
}
```
