---
title: "Summary-statistics Mendelian randomization and GATE score"
subtitle: "PDCD1 protein and type 1 diabetes"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    theme: "Boadilla"
    colortheme: "beaver"
    fonttheme: "structurebold"
    latex_engine: pdflatex
    slide_level: 2
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \setbeamertemplate{navigation symbols}{}
  - \setbeamerfont{frametitle}{size=\normalsize}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center")
library(data.table)
library(ggplot2)
library(ggrepel)
library(knitr)
library(kableExtra)
devtools::load_all(".", quiet = TRUE)
```

```{r load-data, include=FALSE}
genes_coding <- fread("refGene.txt.gz", header = FALSE,
                      select = c(2L, 3L, 5L, 6L, 13L),
                      col.names = c("accession","chrom","txStart","txEnd","symbol"))
genes_coding[, chr_g := suppressWarnings(as.integer(sub("^chr","",chrom)))]
genes_coding <- genes_coding[!is.na(chr_g) & startsWith(accession,"NM_")]
genes_coding <- genes_coding[, .(txStart=min(txStart), txEnd=max(txEnd)),
                               by=.(chr=chr_g, symbol)]

fill_na_genes <- function(dt) {
    na_idx <- which(is.na(dt$gene))
    for (i in na_idx) {
        hits <- genes_coding[chr==as.integer(dt$chr[i]) &
                             ((txStart>=dt$locus_start[i] & txStart<=dt$locus_end[i]) |
                              (txEnd  >=dt$locus_start[i] & txEnd  <=dt$locus_end[i]))]
        if (nrow(hits) > 0L) dt$gene[i] <- paste(unique(hits$symbol), collapse=", ")
    }
    dt
}
find_cis <- function(dt) {
    chr2 <- dt[chr=="2"]
    chr2$qtlname[which.min(abs(chr2$locus_start - 241858000L))]
}
make_coeffs <- function(alpha_rds, gamma_rds, genes_rds) {
    alpha_dt <- readRDS(alpha_rds)
    gamma_dt <- readRDS(gamma_rds)
    extra_cols <- intersect(c("sd.Z","n_ld_blocks","min_maf"), names(alpha_dt))
    alpha_sel  <- c("qtlname","chr","locus_start","locus_end",
                    "n_snps","eff_rank","alpha_hat","se.alpha_hat", extra_cols)
    dt <- merge(alpha_dt[, ..alpha_sel],
                gamma_dt[, .(qtlname, gamma_hat, se.gamma_hat)], by="qtlname")
    dt[, chr := as.character(chr)]
    dt <- get_coeffratios(dt, use.delta=TRUE)
    genes <- readRDS(genes_rds)
    dt <- merge(dt, genes, by="qtlname", all.x=TRUE)
    dt <- fill_na_genes(dt)
    dt[, chr_sort := suppressWarnings(as.integer(chr))]
    setorder(dt, chr_sort, locus_start); dt[, chr_sort := NULL]; dt
}
gene_overrides <- list(
    list(chr=1L,  lo=113e6, hi=115e6, gene="PTPN22"),
    list(chr=2L,  lo=240e6, hi=243e6, gene="PDCD1"),
    list(chr=9L,  lo=  5e6, hi=  6e6, gene="CD274"),
    list(chr=10L, lo=  5e6, hi=  7e6, gene="IL15RA"),
    list(chr=12L, lo=109e6, hi=113e6, gene="SH2B3"),
    list(chr=19L, lo= 48e6, hi= 50e6, gene="FUT2"))
apply_overrides <- function(dt) {
    for (ov in gene_overrides) {
        idx <- which(as.integer(dt$chr)==ov$chr &
                     dt$locus_start>=ov$lo & dt$locus_start<=ov$hi)
        if (length(idx)) set(dt, idx, "gene", ov$gene)
    }; invisible(dt)
}

coeffs_ukb <- make_coeffs("alpha_pdcd1_UKB.rds","gamma_t1d_ukb.rds","locus_genes_ukb.rds")
cis_ukb    <- find_cis(coeffs_ukb)
ivw_ukb_trans <- get_estimatorsMR(coeffs_ukb[qtlname != cis_ukb])
ivw_ukb_trans[, pvalue.formatted := format.z.aspvalue(Estimate/SE)]
apply_overrides(coeffs_ukb)

indiv_raw <- fread("pdcd1.mr.coeffs.csv")
setnames(indiv_raw, c("chrom","startpos","endpos","alpha.hat","se.alpha.hat"),
                    c("chr","locus_start","locus_end","alpha_hat","se.alpha_hat"))
indiv_raw[, chr := as.character(chr)]
indiv_raw[, gene := sub(",.*$","",qtlname)]
indiv_raw[, chr_sort := suppressWarnings(as.integer(chr))]
setorder(indiv_raw, chr_sort, locus_start); indiv_raw[, chr_sort := NULL]
indiv_dt <- get_coeffratios(indiv_raw, use.delta=TRUE)
cis_indiv <- indiv_dt[qtl_type=="cis-x", scoreid]
ivw_indiv_trans <- get_estimatorsMR(indiv_dt[!(scoreid %in% cis_indiv)])
ivw_indiv_trans[, pvalue.formatted := format.z.aspvalue(Estimate/SE)]
apply_overrides(indiv_dt)

comp_direct <- readRDS("gamma_comparison_analytic_vs_direct.rds")
r_ad <- cor(comp_direct$gamma_analytic_sdz, comp_direct$gamma_direct, use="complete.obs")
comp_direct <- merge(comp_direct,
                     coeffs_ukb[, .(qtlname, chr, locus_start, locus_end)],
                     by.x="locus_id", by.y="qtlname", all.x=TRUE)
comp_direct[, chr := as.character(chr)]

coeffs_gs  <- make_coeffs("alpha_pdcd1_UKB.rds","gamma_t1d_gs.rds","locus_genes_gs.rds")
cis_gs     <- find_cis(coeffs_gs)
apply_overrides(coeffs_gs)

## Compute d_sc for Analysis 2 scatter (needs ivw_direct_trans)
d_sc <- merge(comp_direct[, .(locus_id, gene, gamma_direct, se_gamma_direct)],
              coeffs_ukb[, .(qtlname, alpha_hat, se.alpha_hat, sd.Z)],
              by.x="locus_id", by.y="qtlname")
d_sc[, theta_IV    := gamma_direct / (alpha_hat * sd.Z)]
d_sc[, se.theta_IV := se_gamma_direct / (alpha_hat * sd.Z)]
d_sc[, inv.var     := se.theta_IV^-2]
is_cis_direct <- d_sc$locus_id == cis_gs
ivw_direct_trans <- d_sc[!is_cis_direct,
                          .(Estimate=sum(theta_IV*inv.var)/sum(inv.var),
                            SE=sqrt(1/sum(inv.var)))]
ivw_direct_trans[, pvalue.formatted := format.z.aspvalue(Estimate/SE)]

gate_res <- if (file.exists("gate_results.rds")) readRDS("gate_results.rds") else NULL
```

## The challenge: linked SNPs and summary statistics

\vspace{0.3em}

**Most GWAS data exist only as summary statistics** (one coefficient per SNP)

\medskip

- SNPs near each other are correlated — *linkage disequilibrium* (LD)
- A univariate (marginal) coefficient absorbs effects from its neighbours
- For Mendelian randomization we need the *direct* effect of each SNP: the
  multivariate (joint) regression coefficient

\medskip

**Individual-level data are often unavailable or siloed across secure servers**

\medskip
$\Rightarrow$ Can we recover multivariate coefficients from summary statistics alone,
using a reference panel for LD?

## Recovering multivariate coefficients via LD inversion

Let $\hat{\boldsymbol\beta}_u$ be the vector of univariate (marginal) per-SNP
coefficients and $\mathbf{C}_g$ the LD correlation matrix from a reference panel.

\medskip

**Key result:**
$$
\hat{\boldsymbol\beta}_m = \mathbf{D}^{-1}\,\mathbf{C}_g^{-1}\,\mathbf{D}\,\hat{\boldsymbol\beta}_u
$$

where $\mathbf{D} = \text{diag}(\text{per-SNP SDs})$.

\medskip

**In plain language — three steps:**

1. Scale each coefficient to *per-SD-genotype* units
2. Multiply by the **inverse** LD correlation matrix (de-correlate)
3. Scale back to *per-allele* units

\medskip
Near-singular LD (many SNPs in high LD) is handled by a *truncated pseudoinverse*
retaining the $K$ largest eigenvalues of $\mathbf{C}_g$ (effective rank).

## Scalar instruments: one number per locus

Each locus yields a *scalar instrument* $Z_i$ — a weighted sum of an individual's
genotypes across all SNPs in the locus:
$$
Z_i = \frac{\sum_j G_{ij}\,\hat\beta_{m,j}}{\|\hat{\boldsymbol\beta}_m\|}
$$

\medskip

**Instrument-exposure coefficient** $\hat\alpha = \|\hat{\boldsymbol\beta}_m\|$

- Expected change in exposure per 1-unit change in $Z$
- Measures instrument *strength*: how much the locus shifts the protein level

\medskip

**Everything computed from summary statistics** using the UKBB EUR LD matrix
($N = 362{,}063$ samples, 18 million variants, GRCh37)

## Instrument-outcome coefficient $\hat\gamma$

$\hat\gamma$ = log-OR of disease per unit change in $Z$ (for binary outcomes)

\medskip

**From summary statistics** — no individual-level outcome data needed:

- Obtain per-SNP outcome $z$-scores from an outcome GWAS
- Apply the same LD inversion to get multivariate outcome coefficients
- Project onto the scalar instrument $Z$

$$
\hat\gamma = \hat\alpha \cdot
  \frac{\displaystyle\sum_j \hat\sigma_j^2\,\hat\gamma_{u,j}\,\hat\beta_{m,j}}
       {\widehat{\operatorname{Var}}(S)}
$$

\medskip

**Alternative** — direct individual-level regression:
score each individual on $Z_i = \sum_j G_{ij}\hat\beta_{m,j}$, regress disease on $Z_i$

Both methods implemented; direct is the ground truth for validation.

## Standard errors

**For $\hat\alpha$** (instrument-exposure):
$$
\operatorname{SE}(\hat\alpha)
= \sqrt{\frac{\sigma_X^2 - \operatorname{Var}(S)}{N \cdot \operatorname{Var}(Z)}}
$$
Exposure variance $\sigma_X^2$ estimated from per-SNP marginal SEs and sample size.

\medskip

**For $\hat\gamma$** (instrument-outcome, binary):
$$
\operatorname{SE}(\hat\gamma)
= \frac{1}{\sqrt{N_Y \cdot \operatorname{Var}(Z) \cdot p(1-p)}}
$$

- Only needs: outcome $N_Y$, case fraction $p$, $\operatorname{Var}(Z)$ from reference panel
- If reference LD $\neq$ target LD, $\operatorname{Var}(Z)$ is mis-estimated $\Rightarrow$ biased SE

\medskip
*Both SEs computed without access to individual-level data.*

## GATE score: pooling instruments across loci

**GATE** (Genetically Aided Trait Estimation) combines all *trans*-pQTL
instruments into a single composite genetic score:
$$
G_i = \sum_{k\,\in\,\text{trans}} \hat\alpha_k\,Z_{ik}
$$

- Each trans-pQTL $k$ contributes its genetically predicted exposure value
- Standardise to SD = 1, then regress disease on $G^*$
- Tests whether the *overall* genetic evidence supports a causal effect

\medskip

**Intuition:** if $\hat\alpha_k > 0$ at many loci and T1D tracks the predicted
protein level, the composite score should associate with T1D more strongly
than any single instrument.

## GATE coefficient from summary statistics

\vspace{0.3em}

Using the score test and independence of trans-loci, the GATE z-statistic reduces to
the **IVW z-statistic** with weights $(\hat\alpha_k^*)^2$:

$$
z_\text{GATE}
= \sqrt{N p(1{-}p)}\cdot
  \frac{\displaystyle\sum_k (\hat\alpha_k^*)^2\,\hat\theta_k}
       {\sqrt{\displaystyle\sum_k (\hat\alpha_k^*)^2}}
$$

where $\hat\alpha_k^* = \hat\alpha_k\,\text{sd}_{Z,k}$ (per-SD units) and
$\hat\theta_k = \hat\gamma_k^*/\hat\alpha_k^*$ is the per-locus IV ratio.

\medskip

**SE of $\hat\beta_\text{GATE}$ simplifies to $1/\sqrt{Np(1-p)}$** — equivalent
to a single 1-df binary logistic regression.

\medskip

Validated against direct individual-level GATE regression ($z_\text{summary} = 12.2$,
$z_\text{direct} = 9.7$; discrepancy consistent with reference–target LD mismatch).

## Application: PDCD1 protein and type 1 diabetes

::: columns
:::: {.column width="55%"}
**Exposure**

- PDCD1 (PD-1): immune checkpoint protein; regulates T-cell activity
- UK Biobank Olink proteomics, $N = 33{,}902$ EUR participants
- `r nrow(coeffs_ukb)` pQTL loci (MAF $\geq 1\%$, $p < 10^{-6}$, HLA excluded)

\medskip

**Outcome**

- Type 1 diabetes: SDRNT1BIO cases + Generation Scotland controls
- 4,922 cases, 7,452 controls (all individual-level)
::::
:::: {.column width="45%"}
**Three analyses**

\footnotesize
\begin{tabular}{lll}
\toprule
 & $\hat\alpha$ & $\hat\gamma$ \\
\midrule
Analysis 1 & indiv. & indiv. \\
(Zhou et al.) & 1000G LD & 1000G LD \\
\midrule
Analysis 2 & sumstats & direct \\
(new) & UKBB LD & indiv. \\
\midrule
Analysis 3 & sumstats & sumstats \\
(new) & UKBB LD & UKBB LD \\
\bottomrule
\end{tabular}
\normalsize
::::
:::

## Analysis 1 (Zhou et al.): $\hat\gamma$ vs $\hat\alpha$

```{r slide-indiv, fig.width=5.5, fig.height=3.8}
plot_iv_estimates(
    alpha_hat    = indiv_dt$alpha_hat,
    se.alpha_hat = indiv_dt$se.alpha_hat,
    gamma_hat    = indiv_dt$gamma_hat,
    se.gamma_hat = indiv_dt$se.gamma_hat,
    theta        = ivw_indiv_trans$Estimate,
    qtlname      = indiv_dt$gene,
    highlight    = indiv_dt$scoreid %in% cis_indiv
) + labs(
    title = paste0("Analysis 1 — 1000G LD  |  ",
                   nrow(indiv_dt), " instruments"),
    subtitle = paste0("IVW (trans only): θ̂ = ",
                      round(ivw_indiv_trans$Estimate, 3),
                      "  (SE ", round(ivw_indiv_trans$SE, 3),
                      ", p = ", ivw_indiv_trans$pvalue.formatted, ")")
)
```

\vspace{-0.5em}
\footnotesize Red point: PDCD1 cis-pQTL. Red dashed line: IVW slope (trans only).

## Analyses 2 and 3: UKBB 18M EUR LD

::: columns
:::: {.column width="50%"}
```{r slide-direct, fig.width=3.2, fig.height=3.2}
plot_iv_estimates(
    alpha_hat    = d_sc$alpha_hat    * d_sc$sd.Z,
    se.alpha_hat = d_sc$se.alpha_hat * d_sc$sd.Z,
    gamma_hat    = d_sc$gamma_direct,
    se.gamma_hat = d_sc$se_gamma_direct,
    theta        = ivw_direct_trans$Estimate,
    qtlname      = sub(",.*$","", d_sc$gene),
    highlight    = is_cis_direct
) + labs(title="Analysis 2: direct γ̂",
         subtitle=paste0("θ̂ = ", round(ivw_direct_trans$Estimate,3),
                         "  (p = ", ivw_direct_trans$pvalue.formatted, ")"))
```
::::
:::: {.column width="50%"}
```{r slide-ukb, fig.width=3.2, fig.height=3.2}
sdz <- coeffs_ukb$sd.Z
plot_iv_estimates(
    alpha_hat    = coeffs_ukb$alpha_hat    * sdz,
    se.alpha_hat = coeffs_ukb$se.alpha_hat * sdz,
    gamma_hat    = coeffs_ukb$gamma_hat    * sdz,
    se.gamma_hat = coeffs_ukb$se.gamma_hat * sdz,
    theta        = ivw_ukb_trans$Estimate,
    qtlname      = sub(",.*$","", coeffs_ukb$gene),
    highlight    = coeffs_ukb$qtlname == cis_ukb
) + labs(title="Analysis 3: LD-approx γ̂",
         subtitle=paste0("θ̂ = ", round(ivw_ukb_trans$Estimate,3),
                         "  (p = ", ivw_ukb_trans$pvalue.formatted, ")"))
```
::::
:::

\vspace{-0.3em}
\footnotesize Both positive, both significant. UKBB LD reference is much larger (362k vs 503 samples).

## Does the LD reference panel matter for $\hat\alpha$?

```{r slide-alpha-compare, fig.width=5, fig.height=3.5}
dt_indiv <- indiv_dt[,  .(gene=gene, chr, lo=locus_start, hi=locus_end,
                           alpha_indiv=alpha_hat)]
dt_ukb   <- coeffs_ukb[, .(chr, lo=locus_start, hi=locus_end,
                            alpha_ukb=alpha_hat*sd.Z)]
setkeyv(dt_ukb,   c("chr","lo","hi"))
setkeyv(dt_indiv, c("chr","lo","hi"))
alpha_comp <- foverlaps(dt_indiv, dt_ukb, nomatch=0L)
slope_ols  <- coef(lm(alpha_ukb ~ 0 + alpha_indiv, data=alpha_comp))["alpha_indiv"]
alpha_comp[, resid_id := alpha_ukb - alpha_indiv]
alpha_comp[, lab := ifelse(abs(resid_id) > 2*sd(resid_id, na.rm=TRUE), gene, "")]

ggplot(alpha_comp, aes(x=alpha_indiv, y=alpha_ukb)) +
    geom_abline(slope=1,         intercept=0, linetype="dashed", colour="grey50") +
    geom_abline(slope=slope_ols, intercept=0, linetype="dotted", colour="blue") +
    geom_point(size=1.5) +
    geom_text_repel(aes(label=lab), size=2.5, max.overlaps=Inf) +
    labs(x=expression(hat(alpha)["Analysis 1, 1000G"]),
         y=expression(hat(alpha)["Analyses 2/3, UKBB 18M"]),
         title="Instrument-exposure coefficients: 1000G vs UKBB LD") +
    coord_fixed() + theme_bw(base_size=10)
```

\vspace{-0.3em}
\footnotesize Grey dashed: identity. Blue dotted: OLS through origin.
Most loci agree well; a few (PDCD1 cis, ASGR1) change with MAF filter applied.

## Does the LD-approximate $\hat\gamma$ agree with the direct estimate?

```{r slide-gamma-compare, fig.width=5, fig.height=3.5}
comp_direct[, resid_id := gamma_analytic_sdz - gamma_direct]
top6 <- comp_direct[order(-abs(resid_id))][seq_len(min(6L,.N)), gene]
comp_direct[, lab := ifelse(gene %in% top6, gene, "")]
ax <- pretty(range(c(comp_direct$gamma_direct, comp_direct$gamma_analytic_sdz), na.rm=TRUE))

ggplot(comp_direct, aes(x=gamma_direct, y=gamma_analytic_sdz)) +
    geom_hline(yintercept=0, linewidth=0.3) +
    geom_vline(xintercept=0, linewidth=0.3) +
    geom_abline(slope=1, intercept=0, linetype="dashed", colour="grey50") +
    geom_point(size=1.5) +
    geom_text_repel(aes(label=lab), size=2.5, max.overlaps=Inf) +
    scale_x_continuous(breaks=ax, limits=range(ax)) +
    scale_y_continuous(breaks=ax, limits=range(ax)) +
    labs(x=expression(hat(gamma)["direct (individual-level)"]),
         y=expression(hat(gamma)["LD-approximate (summary stats)"]),
         title=paste0("LD-approximate vs direct γ̂:  r = ",
                      round(r_ad, 3))) +
    coord_fixed() + theme_bw(base_size=10)
```

\vspace{-0.3em}
\footnotesize Pearson $r = `r round(r_ad, 3)`$. Attenuation reflects LD mismatch between UKBB reference and GS target.

## GATE score: summary-level vs direct regression

```{r slide-gate, results='asis'}
if (!is.null(gate_res)) {
    tab <- data.frame(
        Method = c("Summary-level (eqs.\\,7--8)", "Direct regression"),
        `$\\hat\\beta$` = sprintf("%.3f", gate_res$beta_GATE),
        SE   = sprintf("%.4f", gate_res$se_GATE),
        z    = sprintf("%.2f",  gate_res$z_GATE),
        p    = formatC(gate_res$p_value, format="g", digits=2),
        `$K$` = as.integer(gate_res$n_trans_pqtl),
        check.names = FALSE
    )
    kable(tab, format="latex", booktabs=TRUE, escape=FALSE,
          col.names=c("Method","$\\hat\\beta$","SE","$z$","$p$","Trans pQTLs")) |>
        kable_styling(font_size=9, latex_options="hold_position")
}
```

\medskip

- Both methods: **strongly positive effect of PDCD1 on T1D**
- $z_\text{direct} = `r if(!is.null(gate_res)) round(gate_res[method=="direct", z_GATE], 1)`$,
  $p = `r if(!is.null(gate_res)) formatC(gate_res[method=="direct", p_value], format="g", digits=2)`$
- $z_\text{summary}/z_\text{direct} \approx 1.26$: reference panel slightly over-estimates
  instrument variance $\Rightarrow$ SE too small by $\sim 26\%$

## Conclusions

\vspace{0.3em}

**Methods**

- Multivariate SNP–exposure and SNP–outcome coefficients can be computed from
  summary statistics by inverting the LD correlation matrix
- SE of $\hat\gamma$ requires only $N$, case fraction, and reference-panel LD
- GATE z-statistic is equivalent to the IVW estimator restricted to trans-pQTLs;
  SE $= 1/\sqrt{Np(1-p)}$ when reference-panel LD is well-calibrated

\medskip

**PDCD1 / T1D results**

- All three analyses consistently estimate a **positive causal effect**
  of circulating PDCD1 on T1D risk
- GATE score confirms this at genome-wide significance ($z \approx 9.7$ direct,
  $z \approx 12.2$ summary-level)

\medskip

**Validation**

- Score tests agree with GLM ($r > 0.9999$, slope $\approx 1$)
- Summary-level $\hat\gamma \approx$ direct ($r = `r round(r_ad, 2)`$)
- Summary-level GATE $\approx$ direct GATE ($z$ ratio 1.26 consistent with known LD mismatch)
